---
title: "Alpha diversity"
---

::: callout-tip
Alpha diversity indices were calculated with Hill numbers (q0 = observed richness, q1 = Shannon Exp and q2 = Inv Simpson). The significance of the observed differences among mangrove systems was determined by employing a linear model, given the imbalanced sample numbers.
:::

------------------------------------------------------------------------

## Load libraries and prepare data

```{r}
#| label: libraries
#| message: false
#| warning: false
#Load libraries
library(phyloseq)
library(hilldiv)
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(cowplot)
library(vegan)
library(emmeans)
library(multcomp)
```

```{r}
#load data
physeq_qiime3 <- readRDS("rds/compare_mangroves/physeq_qiime3.rds")
```

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Términos Lagoon" = "#329D9C",
                "Celestún Lagoon" = "#41e8d3")
```

```{r}
# Extract data from phyloseq object
#library(phyloseq)
otu_data <- otu_table(physeq_qiime3, taxa_are_rows = TRUE)
metadata <- as(sample_data(physeq_qiime3), "data.frame")
```

## 01. Get Hill numbers

```{r}
# Calculate hill numbers
#library(hilldiv)

q0 <- hill_div(otu_data, qvalue = 0)
q1 <- hill_div(otu_data, qvalue = 1)
q2 <- hill_div(otu_data, qvalue = 2)

# Merge metadata with Hill numbers
#library(tidyverse)
q012 <- cbind(q0, q1, q2) %>% as.data.frame() %>%
  rownames_to_column(var = "SampleID")

# join
metadata_with_hill <- q012 %>%
  inner_join(metadata, by = c("SampleID"="SampleID"))

# show
metadata_with_hill %>% head()

#save table
write.table(metadata_with_hill, "Tables/SurfaceComparison/Metadata_with_hill.tsv",
            quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)

```

## 02. Get Hill means

```{r}
# Reorder q values
meta_qs <- metadata_with_hill %>%
  pivot_longer(cols = q0:q2, names_to = "q", values_to = "value") %>%
  filter(q %in% c("q0", "q1", "q2")) %>%
  mutate(
    qs = case_when(
      q == "q0" ~ "q0=Observed",
      q == "q1" ~ "q1=Exp Shannon",
      q == "q2" ~ "q2=Inv Simpson",
    ))

#Get means of Hill numbers
means <- meta_qs %>% group_by(Mangrove_type, Mangrove.system, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')

print(means)

#save table
write.table(means, "Tables/SurfaceComparison/Hill_means_sd.tsv", quote = FALSE,
            sep = "\t", row.names = TRUE, col.names=TRUE)

# group by mangrove type
means_mangrove_type <- meta_qs %>% group_by(Mangrove_type, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')

print(means_mangrove_type)
write.table(means, "Tables/SurfaceComparison/Hill_means_sd_mangrove_type.tsv", quote = FALSE,
            sep = "\t", row.names = TRUE, col.names=TRUE)

```

## 03. Get sequencing effort

```{r}
#| message: false
#| warning: false
### Sample effort
#library(vegan)
mat <- as(t(otu_table(physeq_qiime3)), "matrix")
raremax <- min(rowSums(mat))

# plot
system.time(rarecurve(mat, step = 100, sample = raremax, col = "green4", 
                      label = FALSE))
# save
pdf("Figures/SurfaceComparison/sample_effort.pdf")
system.time(rarecurve(mat, step = 100, sample = raremax, col = "green4", 
                      label = FALSE))
dev.off()

```

## 04. Get significant differences

::: callout-caution
## Remember

The significance of the observed differences among mangrove systems was determined by employing a linear model, given the imbalanced sample numbers.
:::

```{r}
# change variable to factor
metadata_with_hill$Mangrove.system <- 
  factor(metadata_with_hill$Mangrove.system)

# Change to Fossil Lagoon reference
metadata_with_hill$Mangrove.system <- 
  relevel(metadata_with_hill$Mangrove.system,ref = "Fossil Lagoon")

```

### q0 Observed richness

```{r}
# linear model adjust
q0lm <- lm(q0 ~ Mangrove.system, data = metadata_with_hill)
# summary
summary(q0lm)

# check model reliability
shapiro.test(residuals(q0lm))

```

```{r}
# get differences with emmeans
#library(emmeans)
q0_lm_means <- emmeans(q0lm, pairwise ~ Mangrove.system)

# get significance letters
#library(multcomp)
cld_results <- cld(object = q0_lm_means$emmeans, Letters = letters)

# Convertir a data frame
q0_emmeans_df <- as.data.frame(cld_results)
q0_emmeans_df

# Reorder samples
metadata_with_hill$Mangrove.system <- factor(
  metadata_with_hill$Mangrove.system,
  levels = c("Fossil Lagoon", "San Pedro River", 
             "Términos Lagoon", "Celestún Lagoon"))
```

```{r}
# boxplot with significance
q0plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system, y = q0, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q0_emmeans_df, aes(y = emmean, label = .group), vjust = -3,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q0 = Observed", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none", axis.text.x = element_blank())
```

### q1 Diversity Exp Shannon

```{r}
# linear model
q1lm <- lm(q1 ~ Mangrove.system, data = metadata_with_hill)

# summary
summary(q1lm)

#check model reliability
shapiro.test(residuals(q1lm))

# significance
q1_lm_means <- emmeans(q1lm, pairwise ~ Mangrove.system)
cld_results <- cld(object = q1_lm_means$emmeans, Letters = letters)

# Convert to data frame
q1_emmeans_df <- as.data.frame(cld_results)
q1_emmeans_df
```

```{r}
# boxplot
q1plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system , y = q1, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q1_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +   labs(title = "q1 = Exp Shannon", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none", axis.text.x = element_blank())
```

### q2 Dominance Inv Simpson

```{r}
# Model
q2lm <- lm(q2 ~ Mangrove.system, data = metadata_with_hill)

# summary
summary(q2lm)

#check model reliability
shapiro.test(residuals(q2lm))

# get significance
q2_lm_means <- emmeans(q2lm, pairwise ~ Mangrove.system)
cld_results <- cld(object = q2_lm_means$emmeans, Letters = letters)

# Convert to data frame
q2_emmeans_df <- as.data.frame(cld_results)
q2_emmeans_df
```

```{r}
# boxplot
q2plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system, y = q2, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q2_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
            color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q2 = Inv Simpson", y = NULL, x = NULL,
  caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none")
```

### Combine plots

```{r}
#Combine alpha diversity plots
# library(ggplot2)
# library(cowplot)

ytitle <- ggdraw() + draw_label("Effective number of ASVs",
                                angle = 90, size = 14 )
q012plot <- plot_grid(q0plot, q1plot, q2plot,
                      labels = c("A)", "", ""), ncol = 1)
q012y_plot <- plot_grid(ytitle, q012plot, ncol = 2, rel_widths = c(0.05, 1))

# show
q012y_plot

# save
ggsave("Figures/SurfaceComparison/alphadiv_per_loc.pdf",
       q012y_plot, width = 6.3, height = 9.9)
```

## Save rds alpha diversity plot

```{r}
saveRDS(q012y_plot, "rds/compare_mangroves/alpha_diversity_surface_plot.rds")
```
