---
title: "Beta diversity"
---

::: callout-tip
The analysis of beta diversity was conducted using both NMDS and PERMANOVA to identify the factors that influenced the structure of the microbial community.
:::

------------------------------------------------------------------------

## Load libraries and prepare data

```{r}
#| label: libraries
#| message: false
#| warning: false
# load libraries
library(vegan)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(pairwiseAdonis)
library(reshape2)
```

```{r}
#load data
physeq_qiime3 <- readRDS("rds/compare_mangroves/physeq_qiime3.rds")
```

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Términos Lagoon" = "#329D9C",
                "Celestún Lagoon" = "#41e8d3")
```

```{r}
# Extract data from phyloseq object
#library(phyloseq)
otu_data <- otu_table(physeq_qiime3, taxa_are_rows = TRUE)
metadata <- as(sample_data(physeq_qiime3), "data.frame")

sample_data <- data.frame(sample_data(physeq_qiime3))
```

## 01. NMDS

### Get distances and stress

::: callout-note
Bray-Curtis dissimilarity distances are a measure of the dissimilarity between microbial communities based on their species composition.
:::

```{r}
# library(vegan)
# library(dplyr)
# distances
set.seed(123)
bray=vegdist(t(otu_data), method = "bray")
nmds_source_bray = vegan::metaMDS(bray, trymax = 20, k = 2)

# stress
var_stress_nmds_bray <- round(nmds_source_bray$stress, 5)

# score
scores_source_bray= nmds_source_bray %>% vegan::scores()
```

### Plot

```{r}
#| message: false
#| warning: false
# library(ggplot2)
# library(tidyverse)

nmds_plot_bray <- ggplot() +
  geom_point(data=data.frame(scores_source_bray) %>%
               rownames_to_column(var = "SampleID")%>%
               left_join(metadata, by = "SampleID"),
             aes(x=NMDS1, y=NMDS2,  color = Mangrove.system, 
                 shape = Mangrove_type),
             size=5) + theme_linedraw()+
  scale_fill_manual(values = loc_colors)+
  scale_color_manual(values = loc_colors)+
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(colour = "black", size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right",
        legend.box = "vertical")+ theme_classic() +
  labs(title="Bray-Curtis distances") +
  ylab("NMDS2")+xlab("NMDS1")

## stress legend
nmds_plot_bray <- nmds_plot_bray +
  annotate("text", x = Inf, y = -Inf,
           label = paste("Stress:", var_stress_nmds_bray),
           hjust = 1.1, vjust = -1.1, size = 4)

# show plot
nmds_plot_bray

```

## 02. PERMANOVA

::: callout-note
To test the effects of Mangrove type and Mangrove system on the Bray-Curtis dissimilarity, a PERMANOVA analysis was conducted using adonis2.
:::

```{r}
# Extract and transpose OTU table
otu_mat <- t(otu_table(physeq_qiime3))

# PERMANOVA with adonis2
set.seed(123)
adonis_result <- adonis2(otu_mat ~ Mangrove_type * Mangrove.system,
                         data = sample_data,
                         method = "bray",
                         permutations = 999,
                         by = "terms")

# show adonis result
print(adonis_result)
```

### Plot with PERMANOVA results

```{r}
#| message: false
#| warning: false
#add to plot
nmds_plot_bray_perm <- nmds_plot_bray + annotate("text", x = Inf, y = -Inf,
           label = paste("PERMANOVA:",
                         "Mangrove system: R² = 0.206, p = 0.001\n",
                         "Mangrove type: R² = 0.283, p = 0.001\n",
                         "Residual: R² = 0.511"),
           hjust = 1.1, vjust = -7, size = 4)

# Mostrar el gráfico
print(nmds_plot_bray_perm)

#save
ggsave("Figures/SurfaceComparison/beta-diversity.pdf", nmds_plot_bray_perm, width = 7.3, height = 5.5)
```

## 03. Post-hoc adonis

::: callout-note
To test differences between the Mangrove systems a Post-hoc pairwise PERMANOVA was conducted using pairwise.adonis
:::

### Pairwise adonis

```{r}
## Post-hoc Mangrove.system
#library(pairwiseAdonis)
pairwise_adonis <- pairwiseAdonis::pairwise.adonis2(otu_mat ~ 
                  Mangrove.system, data = sample_data, 
                  method = "bray",  permutations = 999)

print(pairwise_adonis)
```

### P-values matrix

```{r}
# heatmap 
# library(ggplot2)
# library(reshape2)

#  p-values of pairwise_adonis
p_values <- c(
  "San Pedro River_vs_Fossil Lagoon" = 0.004,
  "San Pedro River_vs_Celestún Lagoon" = 0.001,
  "San Pedro River_vs_Términos Lagoon" = 0.007,
  "Fossil Lagoon_vs_Celestún Lagoon" = 0.001,
  "Fossil Lagoon_vs_Términos Lagoon" = 0.02,
  "Celestún Lagoon_vs_Términos Lagoon" = 0.003
)

# matrix of p-values
locations <- c("San Pedro River", "Fossil Lagoon", "Celestún Lagoon", "Términos Lagoon")
p_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                   dimnames = list(locations, locations))

# Fill matrix with p-values (only for the upper half then make symmetrical)
pairs <- strsplit(names(p_values), "_vs_")
for (i in seq_along(pairs)) {
  row_name <- pairs[[i]][1]
  col_name <- pairs[[i]][2]
  p_matrix[row_name, col_name] <- p_values[i]
  p_matrix[col_name, row_name] <- p_values[i]  # symmetrical
}
diag(p_matrix) <- 1  # compare with itself

# long format
p_melt <- melt(p_matrix)
```

### Heatmap of adonis posthoc

```{r}
# heatmap
heatmap_plot <- ggplot(p_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0.025, limits = c(0, 0.05),
                       name = "p-value") +
  geom_text(aes(label = sprintf("%.3f", value)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "Pairwise PERMANOVA post-hoc)",
       x = "Mangrove system", y = "Mangrove system") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14))

# show
print(heatmap_plot)
```

## 04. Beta disper

::: callout-note
To validate the dissimilarity of the data, and to determine whether the differences found are due to changes in composition or to variations in dispersion within groups, the homogeneity of the dispersions between Mangrove type and Mangrove system was calculated using betadisper. Subsequently, the significance of the differences in dispersion was calculated with permutest
:::

```{r}
# library(vegan)
# library(phyloseq)
# Bray distance matrix
dist_matrix <- phyloseq::distance(physeq_qiime3, method = "bray")

```

### By Mangrove system

```{r}
# betadisper
disp_location <- betadisper(dist_matrix, sample_data$Mangrove.system)
print(summary(disp_location))
```

Permutest

```{r}
# Significance permutest
perm_location <- permutest(disp_location, permutations = 999)
print(perm_location)
```

```{r}
# Plots
plot(disp_location, main = "Beta Dispersión por Mangrove system")
boxplot(disp_location, main = "Distancias al centroide por Mangrove system")
```

### By Mangrove type

```{r}
# Betadisper Mangrove_type
disp_mangrove_type <- betadisper(dist_matrix, sample_data$Mangrove_type)
print(summary(disp_mangrove_type))

# Significance permutest
perm_mangrove_type <- permutest(disp_mangrove_type, permutations = 999)
print(perm_mangrove_type)
```

```{r}
# plots
plot(disp_mangrove_type, main = "Betadisper  Mangrove_type")
boxplot(disp_mangrove_type, main = "Dist to centroid Mangrove_type")
```

## Save rds beta diversity plot

```{r}
saveRDS(nmds_plot_bray_perm, "rds/compare_mangroves/beta-diversity-surface-plot.rds")
```
