---
title: "Beta diversity"
---

# Load libraries and prepare data

```{r}
#| label: libraries
#| message: false
#| warning: false
#Load libraries
library(vegan)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(pairwiseAdonis)
library(reshape2)
library(ggpubr)
library(cowplot)
```

```{r}
#load data
ps <- readRDS("rds/interior_mangroves/phyloseq.rds")
```

```{r}
# Extract data from phyloseq object
#library(phyloseq)
otu_data <- otu_table(ps, taxa_are_rows = TRUE)
metadata <- as(sample_data(ps), "data.frame")
sample_data <- data.frame(sample_data(ps))
```

```{r}
# color pallet
depth_colors <- c("0-15" = "#f5e5c4",
                 "16-30" = "#baa179",
                 "31-45" = "#a67451",
                 "50-75" = "#80673b")

loc_colors <- c("Fossil Lagoon"= "#A3D9A8",
                "Reforma Waterfalls" = "#44C17B",
                "Miguelito Dike" = "#2E8B57")
```

# 01. Weighted Unifrac

## 01.1 All samples

### 01.1.1 Get distances and stress

```{r}
# NMDS wu
set.seed(123)
nmds_wunifrac <- ordinate(ps, method = "NMDS", distance = "wunifrac")

# stress variable
var_stress_nmds_wu <- round(nmds_wunifrac$stress, 5)
var_stress_nmds_wu
```

### 01.1.2 Plot

```{r}
#| message: false
#| warning: false
# Convert Depth to factor 
sample_data(ps)$Depths <- factor(sample_data(ps)$Depths, levels = c("0-15", "16-30", "31-45", "50-75"))

#plot
nmds_wu <- plot_ordination(ps, nmds_wunifrac, color = "Depths") + theme_classic() + 
  labs(col = "Depth") + 
  labs(title="Weighted UniFrac") +
  geom_point(size=5) +
    scale_fill_manual(values = depth_colors)+ 
    scale_color_manual(values = depth_colors)

nmds_wu <- nmds_wu +
  annotate("text", x = Inf, y = -Inf, label = paste("Stress:", var_stress_nmds_wu),
           hjust = 1.1, vjust = -1.1, size = 4)

#show
nmds_wu
```

### 01.1.3 PERMANOVA

::: callout-note
To test the effects of Depth and Location on the Weighted Unifrac dissimilarity, a PERMANOVA analysis was conducted using adonis2.
:::

```{r}
# UniFrac
wunifrac_distances <- UniFrac(ps, weighted = TRUE)

# Convert distances 'dist' class
wunifrac_distances <- as.dist(wunifrac_distances)

# PERMANOVA 
set.seed(123)
adonis2_wunifrac <- adonis2(
  wunifrac_distances ~ Depths + Location,
  data = metadata,
  permutations = 999,
  strata = metadata$ID,
  by = "terms")

# Results
perm_results <- data.frame(
  Term = c("Depth_cm", "Location", "Residual"),
  R2 = c(adonis2_wunifrac$R2[1], adonis2_wunifrac$R2[2], adonis2_wunifrac$R2[3]),
  p_value = c(adonis2_wunifrac$`Pr(>F)`[1], adonis2_wunifrac$`Pr(>F)`[2], NA))

perm_results

# PERMANOVA text
perm_text <- paste(
  "PERMANOVA:",
  paste("Depths: R² =", round(perm_results$R2[1], 3), ", p =", perm_results$p_value[1]),
  paste("Location: R² =", round(perm_results$R2[2], 3), ", p =", perm_results$p_value[2]),
  paste("Residual: R² =", round(perm_results$R2[3], 3)),
  sep = "\n")
```

### 01.1.4 Plot with permanova

```{r}
#| warning: false

# add to plot
nmds_wu <- nmds_wu +
  annotate(
    "text", x = Inf, y = Inf, label = perm_text,
    hjust = 1.1, vjust = 1.6, size = 3, color = "black"
  )

# Mostrar el gráfico
nmds_wu
```

```{r}
ggsave("Figures/interior_mangroves/nmds_wunifrac_all.pdf", nmds_wu, width = 20, height = 15, units = "cm")
```

## 01.2 Surface and deep samples

### 01.2.1 Filter depths

```{r}
# Filter data 0-15 y 50-75
sample_data(ps)$Depths_fltr <- factor(sample_data(ps)$Depths, levels = c("0-15", "50-75"))

ps_fltr <- prune_samples(sample_data(ps)$Depths_fltr %in% c("0-15", "50-75"), ps)

metadata_fltr <- as(sample_data(ps_fltr), "data.frame")
```

### 01.2.2 Get distances and stress

```{r}
# Distances
set.seed(123)
nmds_wunifrac_fltr <- ordinate(ps_fltr, method = "NMDS", distance = "wunifrac")

# stress
var_stress_nmds_wu_fltr <- round(nmds_wunifrac_fltr$stress, 5)
```

### 01.2.3 NMDS plot

```{r}
# Plot NMDS
nmds_wu_fltr <- plot_ordination(ps_fltr, nmds_wunifrac_fltr, color = "Depths") + 
  theme_classic() + 
  labs(col = "Depths") + 
  labs(title = "Weighted UniFrac") +
  geom_point(size = 5) +
  scale_fill_manual(values = depth_colors) + 
  scale_color_manual(values = depth_colors)

# add stress
nmds_wu_fltr <- nmds_wu_fltr +
  annotate("text", x = Inf, y = -Inf, label = paste("Stress:", var_stress_nmds_wu_fltr),
           hjust = 1.1, vjust = -1.2, size = 3)
```

### 01.2.4 PERMANOVA

```{r}
# UniFrac distances
wunifrac_distances_fltr <- UniFrac(ps_fltr, weighted = TRUE)

# Convert 'dist' class
wunifrac_distances_fltr <- as.dist(wunifrac_distances_fltr)

# PERMANOVA 
set.seed(123)
adonis2_wunifrac_fltr <- adonis2(
  wunifrac_distances_fltr ~ Depths_fltr + Location,
  data = metadata_fltr,
  permutations = 999,
  strata = metadata_fltr$ID,
  by = "terms")

# Results
perm_results_fltr <- data.frame(
  Term = c("Depth_cm", "Location", "Residual"),
  R2 = c(adonis2_wunifrac_fltr$R2[1], adonis2_wunifrac_fltr$R2[2], adonis2_wunifrac_fltr$R2[3]),
  p_value = c(adonis2_wunifrac_fltr$`Pr(>F)`[1], adonis2_wunifrac_fltr$`Pr(>F)`[2], NA))

perm_results_fltr

# PERMANOVA text
perm_text_fltr <- paste(
  "PERMANOVA:",
  paste("Depths: R² =", round(perm_results_fltr$R2[1], 3), ", p =", perm_results_fltr$p_value[1]),
  paste("Location: R² =", round(perm_results_fltr$R2[2], 3), ", p =", perm_results_fltr$p_value[2]), sep = "\n")
```

### 01.2.5 Post-hoc adonis

::: callout-note
To test differences between the Mangrove systems a Post-hoc pairwise PERMANOVA was conducted using pairwise.adonis
:::

### 01.2.6 Pairwise adonis

```{r}
## Post-hoc Depths
# Pairwise PERMANOVA  Weighted UniFrac
set.seed(123)
pairwise_adonis_fltr <- pairwiseAdonis::pairwise.adonis2(
  wunifrac_distances_fltr ~ Depths_fltr,
  data = metadata_fltr,
  permutations = 999)

# show
pairwise_adonis_fltr

# Extract pairwise adonis results
pairwise_text_fltr <- paste(
  "Pairwise PERMANOVA Depths:",
  paste("R² =", round(pairwise_adonis_fltr$`0-15_vs_50-75`$R2[1], 3), 
        ", p =", pairwise_adonis_fltr$`0-15_vs_50-75`$`Pr(>F)`[1]),
  sep = "\n")
```

### 01.2.7 Beta disper

::: callout-important
Given the pairwise PERMANOVA showed a highly significant result (p = 0.001), it is prudent to calculate the beta dispersion to validate that this result reflects real differences in microbial community composition between “0-15” and “50-75,” and not just differences in variability within each depth. In addition, R² = 0.346 indicates a moderate effect, but the 65.4% residual variance suggests unexplained variability, which could include differences in dispersion.
:::

```{r}
# Beta disp
beta_disp_fltr <- betadisper(wunifrac_distances_fltr, metadata_fltr$Depths_fltr)

# Significant test
anova_beta_disp_fltr <- anova(beta_disp_fltr)
anova_beta_disp_fltr

# Show dispersion by group
beta_disp_fltr$group.distances

# boxplot betadisper
beta_data_fltr <- data.frame(
  Group = metadata_fltr$Depths_fltr,
  Distance_to_centroid = beta_disp_fltr$distances
)
beta_plot_fltr <- ggplot(beta_data_fltr, aes(x = Group, y = Distance_to_centroid, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = depth_colors) +
  labs(title = "Betadisper by depth wUniFrac") +
  theme_classic()

#show
beta_plot_fltr
```

Save boxplot

```{r}
ggsave("Figures/interior_mangroves/Betadisper_depth.pdf", beta_plot_fltr, bg='transparent', width = 4, height = 4.5)
```

### 01.2.8 Plot with PERMANOVA, pairwise PERMANOVA and Betadisper

```{r}
# add betadisper p-value text
beta_text_fltr <- paste("Betadisper p-value:", round(anova_beta_disp_fltr$`Pr(>F)`[1], 3))
```

```{r}
#| warning: false
# Añadir texto de PERMANOVA y dispersión beta
nmds_wu_fltr <- nmds_wu_fltr +
  annotate(
    "text", x = Inf, y = Inf, 
    label = paste(perm_text_fltr, pairwise_text_fltr, beta_text_fltr, sep = "\n"),
    hjust = 1.1, vjust = 1.2, size = 3, color = "black")
nmds_wu_fltr
```

save

```{r}
ggsave("Figures/interior_mangroves/NMDS_wu_depth.pdf", nmds_wu_fltr, bg='transparent', width = 6, height = 4.5)
```

# 02. Unweighted UniFrac

## 02.1 Get distances and stress

```{r}
## Unweighted UniFrac
nmds_unifrac <- ordinate(ps, method = "NMDS", distance = "unifrac")
# stress variable
var_stress_nmds_u <- round(nmds_unifrac$stress, 5)
var_stress_nmds_u
```

## 02.2 NMDS

```{r}
# Reorder 'Location'
sample_data(ps)$Location <- factor(sample_data(ps)$Location, levels = c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike"))

# NMDS
nmds_u <- plot_ordination(ps, nmds_unifrac, color = "Location") + 
  theme_classic() + 
  labs(col = "Location", title = "Unweighted UniFrac") +
  geom_point(size = 5) +
  scale_fill_manual(values = loc_colors) + 
  scale_color_manual(values = loc_colors) +
  annotate("text", x = Inf, y = -Inf, 
           label = paste("Stress:", var_stress_nmds_u),
           hjust = 1.1, vjust = -1.3, size = 3)
```

## 02.3 PERMANOVA

```{r}
# UniFrac distances
unifrac_distances <- UniFrac(ps, weighted = FALSE)

# Convert 'dist' class
unifrac_distances <- as.dist(unifrac_distances)

# PERMANOVA 
set.seed(123)
adonis2_unifrac <- adonis2(
  unifrac_distances ~ Location + Depths,
  data = metadata,
  permutations = 999,
  strata = metadata$ID,
  by = "terms")

# Results
perm_results <- data.frame(
  Term = c("Location","Depths", "Residual"),
  R2 = c(adonis2_unifrac$R2[1], adonis2_unifrac$R2[2], adonis2_unifrac$R2[3]),
  p_value = c(adonis2_unifrac$`Pr(>F)`[1], adonis2_unifrac$`Pr(>F)`[2], NA))

perm_results

# PERMANOVA text
perm_text <- paste(
  "PERMANOVA:",
  paste("Location: R² =", round(perm_results$R2[1], 3), ", p =", perm_results$p_value[1]),
  paste("Depths: R² =", round(perm_results$R2[2], 3), ", p =", perm_results$p_value[2]), sep = "\n")
```

## 02.4 Post-hoc adonis

::: callout-important
To test differences between the Mangrove systems a Post-hoc pairwise PERMANOVA was conducted using pairwise.adonis
:::

## 02.5 Pairwise adonis

```{r}
## Post-hoc Location
# Pairwise PERMANOVA  Weighted UniFrac
set.seed(123)
pairwise_adonis_u <- pairwiseAdonis::pairwise.adonis2(
  unifrac_distances ~ Location + Depths,
  data = metadata,
  permutations = 999)

# show
pairwise_adonis_u
```

## 02.6 Extract p-values

```{r}
# Define abbrevs
location_abbrevs <- c(
  "Miguelito Dike" = "MD",
  "Reforma Waterfalls" = "RW",
  "Fossil Lagoon" = "FL")

# Extract values
locations <- unique(metadata$Location)
p_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                   dimnames = list(locations, locations))
r2_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                    dimnames = list(locations, locations))

for (comp in names(pairwise_adonis_u)[-1]) {  # Exclude 'parent_call'
  pair <- strsplit(comp, "_vs_")[[1]]
  loc1 <- pair[1]
  loc2 <- pair[2]
  p_matrix[loc1, loc2] <- pairwise_adonis_u[[comp]]$`Pr(>F)`[1]
  p_matrix[loc2, loc1] <- pairwise_adonis_u[[comp]]$`Pr(>F)`[1]  # Simetric
  r2_matrix[loc1, loc2] <- pairwise_adonis_u[[comp]]$R2[1]
  r2_matrix[loc2, loc1] <- pairwise_adonis_u[[comp]]$R2[1]  # Sim
}
diag(p_matrix) <- 1  # p-value = 1 
diag(r2_matrix) <- 0  # R² = 0 

# abbrevs in matrix
abbrev_locations <- location_abbrevs[locations]
names(abbrev_locations) <- NULL
dimnames(p_matrix) <- list(abbrev_locations, abbrev_locations)
dimnames(r2_matrix) <- list(abbrev_locations, abbrev_locations)

# Convert to long format
p_melt <- melt(p_matrix, na.rm = TRUE)
colnames(p_melt) <- c("Location1", "Location2", "p_value")
r2_melt <- melt(r2_matrix, na.rm = TRUE)
colnames(r2_melt) <- c("Location1", "Location2", "R2")

```

## 02.7 Heatmap

```{r}
# Heatmap of p-values
heatmap_plot_p <- ggplot(p_melt, aes(x = Location1, y = Location2, fill = p_value)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient2(low = "white", mid = "white", high = "gray40",
                       midpoint = 0.025, limits = c(0, 1),
                       name = "p-value") +
  geom_text(aes(label = sprintf("%.3f", p_value)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "Pairwise PERMANOVA Unweighted UniFrac", y = "Location") +
  theme(axis.text.x = element_blank(),  # Ocult axis x
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12),
        plot.title = element_text(size = 12))

# Heatmap de R²
heatmap_plot_r2 <- ggplot(r2_melt, aes(x = Location1, y = Location2, fill = R2)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient(low = "white", high = "gray40",
                      name = "R²") +
  geom_text(aes(label = sprintf("%.3f", R2)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = NULL,
       x = "Location", y = "Location") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, size = 12),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 12))

# join cowplot
combined_heatmap <- plot_grid(
  heatmap_plot_p, heatmap_plot_r2,
  nrow = 2, align = "v", axis = "lr"
)

# show
combined_heatmap
```

save

```{r}
ggsave("Figures/interior_mangroves/pairwiseheatmap.pdf", combined_heatmap, bg='transparent', width = 4, height = 4.5)
```

## 02.8 Beta disper

::: callout-important
Given the pairwise PERMANOVA showed a highly significant results, it is prudent to calculate the beta dispersion to validate that this result reflects real differences in microbial community composition between mangrove location and not just differences in variability within mangrove site.
:::

```{r}
# Beta disp
beta_disp_u <- betadisper(unifrac_distances, metadata$Location)
anova_beta_disp_u <- anova(beta_disp_u)

# Show dispersion by group
beta_disp_u$group.distances

# boxplot betadisper
beta_data <- data.frame(
  Group = metadata$Location,
  Distance_to_centroid = beta_disp_u$distances
)
beta_plot <- ggplot(beta_data, aes(x = Group, y = Distance_to_centroid, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = loc_colors) +
  labs(title = "Betadisper by location wUniFrac") +
  theme_classic()

#show
beta_plot
```

```{r}
ggsave("Figures/interior_mangroves/Betadisper_location.pdf", beta_plot, bg='transparent', width = 5.5, height = 4)
```

## 02.9 Plot with PERMANOVA, pairwise PERMANOVA and Betadisper

```{r}
beta_text <- paste("Betadisper, p-value:", 
                   round(anova_beta_disp_u$`Pr(>F)`[1], 3))
```

```{r}
#| warning: false
# Add texts to NMDS
nmds_u <- nmds_u +
  annotate(
    "text", x = Inf, y = Inf, 
    label = paste(perm_text, beta_text, sep = "\n"),
    hjust = 1.1, vjust = 1.2, size = 3, color = "black"
  )
```

Save

```{r}
ggsave("Figures/interior_mangroves/NMDS_u_location.pdf", nmds_u, bg='transparent', width = 6, height = 4.5)
```

# 03. dbRDA

## 03.1 Load an prepare data

```{r}
# read data
phychem <- read.csv("data/physicochemists.csv", stringsAsFactors = FALSE)

# Rename specific levels
phychem <- phychem %>%
  mutate(`Locality` = case_when(
    Locality == "El Cacahuate" ~ "Fossil Lagoon",
    Locality == "La Piedad" ~ "Reforma Waterfalls",
    Locality == "Dique Miguelito" ~ "Miguelito Dike",
    TRUE ~ Locality))

# extract intersticial samples 30 cm
ps_phychem30cm <- prune_samples(sample_data(ps)$Depths == "31-45", ps)

# Extract metadata from filter object
metadata_phychem30cm <- as.data.frame(sample_data(ps_phychem30cm))

# convert to data frame
metadata_phychem30cm_df <- data.frame(as(sample_data(ps_phychem30cm), 
                                         "data.frame"))
# join
combined_metadata <- merge(metadata_phychem30cm_df, phychem,
                           by = "Sample", all.x = TRUE)

# rownames
rownames(combined_metadata) <- combined_metadata$Sample

# Convert to sample_data
sample_data_combined <- sample_data(combined_metadata)

# save into phyloseq object
sample_data(ps_phychem30cm) <- sample_data_combined
```

## 03.2 Get distances

```{r}
# Calcular las distancias Unifrac no ponderadas
unifrac_distances_phychem30cm<- distance(ps_phychem30cm, method = "unifrac", weighted = FALSE)
```

## 03.3 Calculate dbRDA

```{r}
# Convert distance matrix to vegdist object
unifrac_vegdist_phychem30cm <- as.dist(unifrac_distances_phychem30cm)

# db-RDA
dbrda_result <- capscale(unifrac_vegdist_phychem30cm ~ Temperature...C. + 
                           Salinity..ppm. + pH + Redox.mV..Eh. + 
                           S.2..mg.l. + SO....mg.l., data = phychem)

# summary
summary(dbrda_result)
```

## 03.4 Get significance

```{r}
# Significance
anova_axes <- anova(dbrda_result, by = "axis", 
                    permutations = 999)
anova_axes

# Significance by variable
anova_terms <- anova(dbrda_result, by = "terms", 
                     permutations = 999)

anova_terms
```

```{r}
# variance by each axis
summary(dbrda_result)$cont$importance
```

## 03.5 PERMANOVA

```{r}
# PERMANOVA
set.seed(123)
permanova_result <- adonis2(unifrac_distances_phychem30cm ~ Temperature...C. + Salinity..ppm. + pH +
Redox.mV..Eh. + S.2..mg.l. + SO....mg.l., 
data = phychem, by="terms", permutations = 999)

# show
permanova_result
```

## 03.6 PERMANOVA Plot

```{r}
# Extract PERMANOVA results
permanova_df <- data.frame(
  Variable = rownames(permanova_result)[1:(nrow(permanova_result)-2)],
  R2 = permanova_result$R2[1:(nrow(permanova_result)-2)],
  p_value = permanova_result$`Pr(>F)`[1:(nrow(permanova_result)-2)])

# names
var_abbrevs <- c(
  "Temperature...C." = "Temperature",
  "Salinity..ppm." = "Salinity",
  "pH" = "pH",
  "Redox.mV..Eh." = "Redox",
  "S.2..mg.l." = "S2",
  "SO....mg.l." = "Sulfates")

permanova_df$Variable <- var_abbrevs[permanova_df$Variable]
```

```{r}
# plot
permanova_plot <- ggplot(permanova_df, aes(x = Variable, 
                        y = R2, fill = p_value < 0.05)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "cyan4", "FALSE" = "gray80"), 
                labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05")) +
  labs(x = "Environmental variable", 
       y = "R² Proportion of Variance Explained", 
       fill = "Significance") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 12))

# show
permanova_plot
```

```{r}
ggsave("Figures/interior_mangroves/Physicochemyc_Significance.pdf", permanova_plot, bg='transparent', width = 5.7, height = 4)
```

## 03.7 dbRDA Plot

```{r}
#| warning: false
# Extract site scores
scores_samples <- scores(dbrda_result, display = "sites")

# Convert to data.frame
scores_df <- data.frame(scores_samples, Location = phychem$Locality)

# Extra environmental scores
env_scores <- scores(dbrda_result, display = "bp")
env_scores_df <- data.frame(env_scores)
env_scores_df$Variable <- rownames(env_scores_df)

# Explain variance percentage
explained_var <- summary(dbrda_result)$cont$importance[2, 1:2] * 100  # CAP1 and CAP2

#plot
capphychem_plot <- ggplot(scores_df, aes(x = CAP1, y = CAP2, 
  color = Location)) + geom_point(size = 5) + theme_classic() +  
  labs(
    col = "Location",
    title = "db-RDA Unweighted UniFrac",
    x = paste0("CAP1 (", round(explained_var[1], 1), "%)"),
    y = paste0("CAP2 (", round(explained_var[2], 1), "%)")
  ) +
  scale_color_manual(values = loc_colors) +  
  geom_segment(data = env_scores_df, aes(
    x = 0, y = 0, xend = CAP1, yend = CAP2),
               arrow = arrow(length = unit(0.15, "cm")), 
    color = "gray20", size = 0.6) +
  # environmental variables
  geom_text(data = env_scores_df, aes(x = CAP1 * 1.1, 
                            y = CAP2 * 1.1, label = Variable),
            color = "black", size = 3.5) +
  theme(legend.position = "right")

capphychem_plot
```

```{r}
#save
ggsave("Figures/interior_mangroves/dbRDA.pdf", capphychem_plot, width = 6, height = 4)
```

# Save rds plots

```{r}
#| warning: false
# save plots
nmds_u
nmds_wu_fltr

saveRDS(nmds_u, "rds/interior_mangroves/nmds_u.rds")
saveRDS(nmds_wu_fltr, "rds/interior_mangroves/nmds_wu.rds")
saveRDS(capphychem_plot, "rds/interior_mangroves/dbRDA.rds")
```
