---
title: "Differential Abundance"
---

# Load libraries and prepare data

```{r}
#| message: false
#| warning: false
# load libraries
library(phyloseq)
library(dplyr)
library(ANCOMBC)
library(tidyverse)
library(cowplot)
library(DT)
options(DT.options = list(
  initComplete = JS("function(settings, json) {",
                    "$(this.api().table().header()).css({'background-color':
  '#000', 'color': '#fff'});","}")))

```

```{r}
#load data
ps <- readRDS("rds/interior_mangroves/phyloseq.rds")
```

# **Differential abundance ANCOMBC2**

## 01. By Depth

### **01.1 Prepare data**

```{r}
#| message: false
#| warning: false
# phyloseq to TreeSummarizedExperiment
tse <- mia::convertFromPhyloseq(ps)

# contrast matrix
contrast_matrices <- list(
  matrix(c(1, -1, 0,
           0, 1, -1,
           0, 0, 1),
         nrow = 3, byrow = TRUE),
  matrix(c(1, 0, -1,
           0, 1, -1,
           0, 0, 1),
         nrow = 3, byrow = TRUE),
  matrix(c(1, 0, 0,
           0, 1, -1,
           0, 0, 1),
         nrow = 3, byrow = TRUE),
  matrix(c(1, -1, 0,
           0, 1, 0,
           0, 0, 1),
         nrow = 3, byrow = TRUE))

# nodes
nodes <- list(1, 1, 1, 1)
```

### 01.2 Run ANCOM-BC2

**Nota para Yaxche y después la quitamos.**

**Corre el siguiente bloque, guarda el resultado en un rds y luego leelo para los filtros. Desactiva el bloque para que no corra mientras se renderiza porque tarda mucho este analisis y al renderizar cada vez será una eternidad.**

``` r
# ANCOMBC2
output <- ancombc2(
  data = tse,
  assay_name = "counts",
  tax_level = "Phylum",
  fix_formula = "Depth_cm + Location",
  rand_formula = "(1 | ID)",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0.10,
  lib_cut = 1000,
  s0_perc = 0.05,
  group = "Depth_cm",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = TRUE,
  trend = TRUE,
  iter_control = list(tol = 1e-2, max_iter = 20, verbose = TRUE),
  em_control = list(tol = 1e-5, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(
    contrast = contrast_matrices,
    node = nodes,
    solver = "ECOS",
    B = 100))
```

```{r}
#| include: false
output <- readRDS("rds/interior_mangroves/output_depth.rds")
```

```{r}
#See results
res_pair <- output$res_pair

head(res_pair)
```

### 01.3 Get differential data

```{r}
# Create dataframe df_fig_pair1 to log-fold change (LFC) values
df_fig_pair1 <- res_pair %>%
  dplyr::filter(`diff_Depth_cm16-30` == 1 |
                  `diff_Depth_cm31-45` == 1 |
                  `diff_Depth_cm50-75` == 1 |
                  `diff_Depth_cm31-45_Depth_cm16-30` == 1 |
                  `diff_Depth_cm50-75_Depth_cm16-30` == 1 |
                  `diff_Depth_cm50-75_Depth_cm31-45` == 1) %>%
  dplyr::mutate(lfc1 = ifelse(`diff_Depth_cm16-30` == 1, round(`lfc_Depth_cm16-30`, 2), NA),
                lfc2 = ifelse(`diff_Depth_cm31-45` == 1, round(`lfc_Depth_cm31-45`, 2), NA),
                lfc3 = ifelse(`diff_Depth_cm50-75` == 1, round(`lfc_Depth_cm50-75`, 2), NA),
                lfc4 = ifelse(`diff_Depth_cm31-45_Depth_cm16-30` == 1, 
                      round(`lfc_Depth_cm31-45_Depth_cm16-30`, 2), NA),
                lfc5 = ifelse(`diff_Depth_cm50-75_Depth_cm16-30` == 1, 
                      round(`lfc_Depth_cm50-75_Depth_cm16-30`, 2), NA),
                lfc6 = ifelse(`diff_Depth_cm50-75_Depth_cm31-45` == 1, 
                      round(`lfc_Depth_cm50-75_Depth_cm31-45`, 2), NA)) %>%
  tidyr::pivot_longer(cols = lfc1:lfc6, names_to = "group", 
                      values_to = "value") %>% dplyr::arrange(taxon)
```

```{r}
# Create dataframe df_fig_pair2 to significant values

df_fig_pair2 <- res_pair %>%
  dplyr::filter(`diff_Depth_cm16-30` == 1 |
                  `diff_Depth_cm31-45` == 1 |
                  `diff_Depth_cm50-75` == 1 |
                  `diff_Depth_cm31-45_Depth_cm16-30` == 1 |
                  `diff_Depth_cm50-75_Depth_cm16-30` == 1 |
                  `diff_Depth_cm50-75_Depth_cm31-45` == 1) %>%
  dplyr::mutate(lfc1 = ifelse(`passed_ss_Depth_cm16-30` == 1 & `diff_Depth_cm16-30` == 1, "darkblue", "black"),
                lfc2 = ifelse(`passed_ss_Depth_cm31-45` == 1 & `diff_Depth_cm31-45` == 1, "darkblue", "black"),
                lfc3 = ifelse(`passed_ss_Depth_cm50-75` == 1 & `diff_Depth_cm50-75` == 1, "darkblue", "black"),
                lfc4 = ifelse(`passed_ss_Depth_cm31-45_Depth_cm16-30` == 1 & `diff_Depth_cm31-45_Depth_cm16-30` == 1, "darkblue", "black"),
                lfc5 = ifelse(`passed_ss_Depth_cm50-75_Depth_cm16-30` == 1 & `diff_Depth_cm50-75_Depth_cm16-30` == 1, "darkblue", "black"),
                lfc6 = ifelse(`passed_ss_Depth_cm50-75_Depth_cm31-45` == 1 & `diff_Depth_cm50-75_Depth_cm31-45` == 1, "darkblue", "black")) %>%
  tidyr::pivot_longer(cols = lfc1:lfc6, names_to = "group", values_to = "color") %>%
  dplyr::arrange(taxon)
```

```{r}
# Combine dataframes df_fig_pair1 and df_fig_pair2
df_fig_pair <- df_fig_pair1 %>%
  dplyr::left_join(df_fig_pair2, by = c("taxon", "group"))

# Rename groups
df_fig_pair$group <- recode(df_fig_pair$group,
                            `lfc1` = "16-30 vs 0-15",
                            `lfc2` = "31-45 vs 0-15",
                            `lfc3` = "50-75 vs 0-15",
                            `lfc4` = "31-45 vs 16-30",
                            `lfc5` = "50-75 vs 16-30",
                            `lfc6` = "50-75 vs 31-45")

df_fig_pair$group <- factor(df_fig_pair$group,
                            levels = c("16-30 vs 0-15",
                                       "31-45 vs 0-15",
                                       "50-75 vs 0-15",
                                       "31-45 vs 16-30",
                                       "50-75 vs 16-30",
                                       "50-75 vs 31-45"))

# Filter comparisons with significant values
df_fig_pair <- df_fig_pair %>%
  group_by(group) %>%
  filter(any(!is.na(value))) %>%
  ungroup()

# Calculate limits
lo <- floor(min(df_fig_pair$value, na.rm = TRUE))
up <- ceiling(max(df_fig_pair$value, na.rm = TRUE))
mid <- (lo + up) / 2
```

### 01.4 Heatmap Plot

```{r}
fig_pair <- df_fig_pair %>%
  ggplot(aes(x = group, y = taxon, fill = value)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient2(low = "#c05098", high = "#3FB6AFFF", mid = "gray90",
                       na.value = "white", midpoint = mid, 
                       limit = c(lo, up),name = "LFC") +
  geom_text(aes(label = round(value, 2), color = color), 
            size = 3.5, na.rm = TRUE) + 
  scale_color_identity(guide = FALSE) +
  labs(x = NULL, y = NULL) + theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.text.x = element_text(size = 10, color = "black"))

fig_pair
```

### 01.5 Relative abundance of differential Phyla

#### 01.5.1 Filter and prepare data

```{r}
#| message: false
#| warning: false

# Filter da phyla
differential_phyla <- c("Zixibacteria", "WS2", "uncultured",
                        "Thermoplasmatota", "TA06", "Sva0485",
                        "Spirochaetota", "RCP2-54", "Nitrospinota", 
                        "NB1-j", "Myxococcota", "Modulibacteria", 
                        "Methylomirabilota", "MBNT15", "LCP-89",
                        "Latescibacterota", "Fibrobacterota", 
                        "Crenarchaeota", "Chloroflexi", 
                        "Calditrichota", "Armatimonadota")

# Filter in phyloseq
ps_da <- subset_taxa(ps, Phylum %in% differential_phyla)

# Depth aggregation
ps_da_agg <- merge_samples(ps_da, "Depths")

# relative abundance
ps_da_agg_rel <- transform_sample_counts(ps_da_agg, function(x) x / sum(x))

# check value and names Depths
depths <- unique(sample_data(ps)$Depths)
sample_names(ps_da_agg_rel) <- depths
sample_data(ps_da_agg_rel)$Depths <- factor(depths, levels = depths)

# Convert to data frame
df_abundance <- psmelt(ps_da_agg_rel)

# summary by depth and phylum
df_abundance_summary <- df_abundance %>%
  group_by(Phylum, Depths) %>%
  summarise(Abundance = mean(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Replace NA with 0
df_abundance_summary$Abundance[is.na(df_abundance_summary$Abundance)] <- 0
```

#### 01.5.2 Bubble plot

```{r}
# Define colors
depth_colors <- c("0-15" = "#f5e5c4",
                  "16-30" = "#baa179",
                  "31-45" = "#a67451",
                  "50-75" = "#80673b")

# bubble plot
bubble_plot <- ggplot(df_abundance_summary, 
                      aes(x = Depths, y = Phylum, size = Abundance,
                          fill = Depths)) +
  geom_point(shape = 21, color = "gray40") +
  scale_size_continuous(range = c(1, 12), guide = "none") +
  scale_fill_manual(values = depth_colors) +
  labs(y = NULL, x=NULL , fill = NULL) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, 
                               size = 8.5, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    legend.position = "none")

# show
bubble_plot
```

### 01.6 Join relative and differential abundance plots

```{r}
fig_pair <- fig_pair +
  theme(
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 8.5 , color = "black"),
    axis.line.y = element_blank(),
    axis.line.x = element_line(color = "black"))
```

```{r}
depth_abundances <- plot_grid(bubble_plot, fig_pair, ncol = 2, rel_widths = c(0.95, 1.1))

depth_abundances
```

```{r}
# save
ggsave("Figures/interior_mangroves/DA_and_RA_BubbleHeatmap_depth.pdf",
       depth_abundances, bg='transparent', 
       width = 6.7, height = 8)
```

### 01.7 Save RDS plot

```{r}
saveRDS(depth_abundances, "rds/interior_mangroves/da_ra_depth_plot.rds")
```

## 02. By Mangrove site

### **02.1 Prepare data**

```{r}
#| message: false
#| warning: false
#To manually change the reference level, not alphabdetic order
tse$Location = factor(tse$Location, levels = c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike"))


#variable
tse$Location = recode(as.character(tse$Location),
                     `Fossil Lagoon` = "FL",
                     `Reforma Waterfalls` = "RW",
                     `Miguelito Dike` = "MD",
                     .missing = "unknown")
```

### 02.2 Run ANCOM-BC2

**Nota para Yaxche y después la quitamos.**

**Corre el siguiente bloque, guarda el resultado en un rds y luego leelo para los filtros. Desactiva el bloque para que no corra mientras se renderiza porque tarda mucho este analisis y al renderizar cada vez será una eternidad.**

``` r
output_location <- ancombc2(
  data = tse,
  assay_name = "counts",
  tax_level = "Phylum",
  fix_formula = "Location + Depth",
  rand_formula = "(1 | ID)",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0.10,
  lib_cut = 1000,
  s0_perc = 0.05,
  group = "Location",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = TRUE,
  trend = TRUE,
  iter_control = list(tol = 1e-2, max_iter = 20, verbose = TRUE),
  em_control = list(tol = 1e-5, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                              nrow = 2,
                                              byrow = TRUE),
                                       matrix(c(-1, 0, 1, -1),
                                              nrow = 2,
                                              byrow = TRUE),
                                       matrix(c(1, 0, 1, -1),
                                              nrow = 2,
                                              byrow = TRUE)),
                       node = list(2, 2, 1),
                       solver = "ECOS",
                       B = 100))
```

``` r
saveRDS(output_location, "rds/interior_mangroves/output_location.rds")
```

```{r}
#| include: false
output_location <- readRDS("rds/interior_mangroves/output_location.rds")
```

``` R
#See results
res_pair_location <- output_location$res_pair

colnames(res_pair_location)
```

### 02.3 Get differential data

``` R
# Create dataframe df_fig_pair1 to log-fold change (LFC) values
df_fig_pair1_location <- res_pair_location %>%
  dplyr::filter(`diff_LocationMD` == 1 |
                  `diff_LocationRW` == 1 |
                  `diff_LocationRW_LocationMD` == 1) %>%
  dplyr::mutate(lfc1 = ifelse(`diff_LocationMD` == 1, round(`lfc_LocationMD`, 2), NA),
                lfc2 = ifelse(`diff_LocationRW` == 1, round(`lfc_LocationRW`, 2), NA),
                lfc3 = ifelse(`diff_LocationRW_LocationMD` == 1, round(`lfc_LocationRW_LocationMD`, 2), NA)) %>%
  tidyr::pivot_longer(cols = lfc1:lfc3, names_to = "group", values_to = "value") %>%
  dplyr::arrange(taxon)
```

``` R
# Create dataframe df_fig_pair2 to significant values
df_fig_pair2_location <- res_pair_location %>%
  dplyr::filter(`diff_LocationMD` == 1 |
                  `diff_LocationRW` == 1 |
                  `diff_LocationRW_LocationMD` == 1) %>%
  dplyr::mutate(lfc1 = ifelse(`passed_ss_LocationMD` == 1 & `diff_LocationMD` == 1, "darkblue", "black"),
                lfc2 = ifelse(`passed_ss_LocationRW` == 1 & `diff_LocationRW` == 1, "darkblue", "black"),
                lfc3 = ifelse(`passed_ss_LocationRW_LocationMD` == 1 & `diff_LocationRW_LocationMD` == 1, "darkblue", "black")) %>%
  tidyr::pivot_longer(cols = lfc1:lfc3, names_to = "group", values_to = "color") %>%
  dplyr::arrange(taxon)
```

::: callout-warning
## Important

The results indicated that there were no significant differences in the differential abundance of Phyla between mangroves along the san pedro river and therefore it is not included in the final results. however, if after the significance filters the results would have shown any differential phyla the subsequent code would be as follows
:::

``` R
#| message: false
#| warning: false
# Combine dataframes df_fig_pair1 and df_fig_pair2
df_fig_pair_location <- df_fig_pair1_location %>%
  dplyr::left_join(df_fig_pair2_location, by = c("taxon", "group"))

# Rename groups
df_fig_pair_location$group <- recode(df_fig_pair_location$group,
                                     `lfc1` = "MD vs FL",
                                     `lfc2` = "RW vs FL",
                                     `lfc3` = "RW vs MD")

df_fig_pair_location$group <- factor(df_fig_pair_location$group,
                                     levels = c("MD vs FL",
                                                "RW vs FL",
                                                "RW vs DM"))


# Filter comparisons with significant values
df_fig_pair_location <- df_fig_pair_location %>%
  group_by(group) %>%
  filter(any(!is.na(value))) %>%
  ungroup()

# Calculate limits
lo <- floor(min(df_fig_pair_location$value, na.rm = TRUE))
up <- ceiling(max(df_fig_pair_location$value, na.rm = TRUE))
mid <- (lo + up) / 2
```

### Heatmap Plot

``` R
fig_pair_location <- df_fig_pair_location %>%
  ggplot(aes(x = group, y = taxon, fill = value)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient2(low = "lightgreen", high = "red4", mid = "gray90",
                       na.value = "white", midpoint = mid, 
                       limit = c(lo, up),name = "LFC") +
  geom_text(aes(label = round(value, 2), color = color), 
            size = 3.5, na.rm = TRUE) + 
  scale_color_identity(guide = FALSE) +
  labs(x = NULL, y = NULL) + theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.text.x = element_text(size = 10, color = "black"))
```
