---
title: "Fig. 3 Interior Mangroves"
---

# Load libraries

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(cowplot)
```

# Load figures

```{r}
#| message: false
#| warning: false

alpha <- readRDS("rds/interior_mangroves/alpha_diversity_interior_depth_plot.rds")

beta_u <- readRDS("rds/interior_mangroves/nmds_u.rds")
beta_wu <- readRDS("rds/interior_mangroves/nmds_wu.rds")
dbrda <- readRDS("rds/interior_mangroves/dbRDA.rds")

core <- readRDS("rds/interior_mangroves/upset_plot.rds")

da_tax <- readRDS("rds/interior_mangroves/da_ra_depth_plot.rds")

da_ko <- readRDS("rds/interior_mangroves/DA_picrust.rds")
```

# Combine plots

```{r}
#| message: false
#| warning: false
alpha_betas <- plot_grid(alpha,beta_wu, beta_u, dbrda, 
                   labels = c("A)","C)","D)","E)"),
                   label_fontface = "bold",
                   ncol = 4, rel_widths = c(0.15,0.25, 0.3, 0.3))
```

```{r}
#| message: false
#| warning: false
dataxko <- plot_grid(da_tax, da_ko, core,
                     labels = c("B)","F)", "G)"),
                     label_fontface = "bold",
                     ncol = 3, rel_widths = c(0.4,0.3, 0.3))
```

```{r}
#| message: false
#| warning: false
all <- plot_grid(alpha_betas, dataxko, 
                 ncol = 1, 
                 rel_heights = c(0.35,0.65))
```

```{r}
ggsave("Figures/interior_mangroves/Fig3.InteriorMangroves.pdf",
       all, width = 20, height = 13)
```

Core and KO differential abundance are not shown because they are not a direct format of ggplot, so we will insert them in Inkscape

::: callout-caution
## Edit figure in Inkscape

The UpSet and ko abundance plots insertion and other details were done in Inkscape. The final figure is shown below:
:::

![](Figures/interior_mangroves/Fig3.InteriorMangroves_InkscapeEdited.pdf){fig-align="center" width="650" height="520"}
