---
title: "Fig. 3 Interior Mangroves"
---

# Load libraries

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(cowplot)
```

# Load figures

```{r}
#| message: false
#| warning: false

alpha <- readRDS("rds/interior_mangroves/alpha_diversity_interior_depth_plot.rds")

beta_u <- readRDS("rds/interior_mangroves/nmds_u.rds")
beta_wu <- readRDS("rds/interior_mangroves/nmds_wu.rds")
dbrda <- readRDS("rds/interior_mangroves/dbRDA.rds")

core <- readRDS("rds/interior_mangroves/upset_plot.rds")

da_tax <- readRDS("rds/interior_mangroves/da_ra_depth_plot.rds")

da_ko <- readRDS("rds/interior_mangroves/DA_picrust.rds")
```

# Combine plots

```{r}
#| message: false
#| warning: false
betas_u_core <- plot_grid(beta_u, dbrda, core, labels = c("A","B", "C"),
                        label_fontface = "bold",
                        ncol = 1, rel_heights = c(0.3, 0.3, 0.5))
```

```{r}
#| message: false
#| warning: false
dataxbeta_wu <- plot_grid(beta_wu,da_tax, labels = c("D","F"),
                        label_fontface = "bold",
                        ncol = 1, rel_widths = c(0.3,0.7))
```

```{r}
#| message: false 
#| warning: false
alpha_kos <- plot_grid(alpha, da_ko,labels = c("E","G"),
                        label_fontface = "bold",
                       ncol = 1, rel_widths = c(0.3,7))
```

```{r}
#| message: false
#| warning: false
all <- plot_grid(betas_u_core, dataxbeta_wu, alpha_kos, 
                 ncol = 3, 
                 rel_widths = c(0.4,0.35,0.30))
```

```{r}
ggsave("Figures/interior_mangroves/Fig3.InteriorMangroves.pdf",
       all, width = 19, height = 13)
```

Core and KO differential abundance are not shown because they are not a direct format of ggplot, so we will insert them in Inkscape

::: callout-caution
## Edit figure in Inkscape

The UpSet and ko abundance plots insertion and other details were done in Inkscape. The final figure is shown below:
:::

![](Figures/interior_mangroves/Fig3.InteriorMangroves_InkscapeEdited.pdf){fig-align="center" width="650" height="520"}
