---
title: "Alpha Diversity"
---

::: callout-note
Alpha diversity indices were calculated with Hill numbers (q0 = observed richness, q1 = Shannon Exp and q2 = Inv Simpson) using *hilldiv* library. The significance of the observed differences among depth and mangrove systems was determined by employing a linear model, given the imbalanced sample numbers.
:::

## Load libraries and prepare data

```{r}
#Load libraries
library(phyloseq)
library(hilldiv)
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(cowplot)
library(vegan)
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
```

```{r}

#load data
physeq_qiime3 <- readRDS("rds/Depth_comparison/physeq_qiime3.rds")
```

```{r}
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Términos Lagoon" = "#329D9C")

depth_colors <- c("Surface" = "#f5e5c4",
                 "mid" = "#baa179",
                 "Deep" = "#80673b")
```

```{r}
# Extract metadata and otu table from phyloseq object

otu_data <- otu_table(physeq_qiime3, taxa_are_rows = TRUE)
metadata <- as(sample_data(physeq_qiime3), "data.frame")
```

## 01.Get Hill numbers

```{r}
# Calculate hill numbers

q0 <- hill_div(otu_data, qvalue = 0)
q1 <- hill_div(otu_data, qvalue = 1)
q2 <- hill_div(otu_data, qvalue = 2)

# Merge metadata with Hill numbers
#library(tidyverse)
q012 <- cbind(q0, q1, q2) %>% as.data.frame() %>%
  rownames_to_column(var = "SampleID")

# join
metadata_with_hill <- q012 %>%
  inner_join(metadata, by = c("SampleID"="SampleID"))

# show
metadata_with_hill %>% head()

#save table
write.table(metadata_with_hill, "Tables/Depth_comparison/Metadata_with_hill.tsv",
            quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)
```

## 02.Get Hill means

```{r}
# Reorder q values
meta_qs <- metadata_with_hill %>%
  pivot_longer(cols = q0:q2, names_to = "q", values_to = "value") %>%
  filter(q %in% c("q0", "q1", "q2")) %>%
  mutate(
    qs = case_when(
      q == "q0" ~ "q0=Observed",
      q == "q1" ~ "q1=Exp Shannon",
      q == "q2" ~ "q2=Inv Simpson",
    ))

#Get means of Hill numbers
means <- meta_qs %>% group_by(Depths, Mangrove.system, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')

print(means)

#save table
write.table(means, "Tables/Depth_comparison/Hill_means_sd.tsv", quote = FALSE,
            sep = "\t", row.names = TRUE, col.names=TRUE)

# group by mangrove system
means_mangrove_system_Depths <- meta_qs %>% group_by(Mangrove.system, Depths, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')

print(means_mangrove_system_Depths)
write.table(means, "Tables/Depth_comparison/Hill_means_sd_mangrove_systemDepths.tsv", quote = FALSE,
            sep = "\t", row.names = TRUE, col.names=TRUE)

```

## 03.Get sequencing effort

```{r}
### Sample effort

mat <- as(t(otu_table(physeq_qiime3)), "matrix")
raremax <- min(rowSums(mat))

# plot
system.time(rarecurve(mat, step = 100, sample = raremax, col = "green4", 
                      label = FALSE))
# save
pdf("Figures/Depth_comparison/sample_effort.pdf")
system.time(rarecurve(mat, step = 100, sample = raremax, col = "green4", 
                      label = FALSE))
dev.off()
```

## 04. Check Alpha diversity correlation with sequencing depth

```{r}
#Get effective reads
Effective_reads <- colSums(otu_data[-1, ])
metadata_with_hill$Effective_reads <- Effective_reads
```

### Alpha diversity depth correlation to q0 order

```{r}
q0_vs_depth <- ggscatter(metadata_with_hill,
                x = "Effective_reads", y = "q0", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=0", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient. 
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")
   )+
  theme(legend.title = element_blank(), legend.position = "none")
```

### Alpha diversity depth correlation to q1 order

```{r}
q1_vs_depth <- ggscatter(metadata_with_hill, 
                x = "Effective_reads", y = "q1", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=1", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")) +
  theme(legend.title = element_blank(), 
        legend.position = "none")
```

### Alpha diversity depth correlation to q2 order

```{r}
q2_vs_depth <- ggscatter(metadata_with_hill, 
                x = "Effective_reads", y = "q2", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=2", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")) +
  theme(legend.title = element_blank(), 
        legend.position = "none")
```

```{r}
#Combine plot
title_corr_plot <- ggdraw() + 
  draw_label("Alpha diversity depth correlation")

correlation_plot_q012 <- plot_grid(title_corr_plot,
                            q0_vs_depth, 
                            q1_vs_depth, 
                            q2_vs_depth,
                            ncol = 1, 
                            rel_heights = c(0.15, 1, 1, 1))

correlation_plot_q012

# #save plot
ggsave("Figures/Depth_comparison/alpha_depth_correlation.pdf", correlation_plot_q012, width = 8, height = 8)
```

## 05.Get significant differences for mangrove system

::: callout-important
Due to the positive correlation of the hill numbers and sequencing depth, we decided to include this variable in the linear model we used.
:::

```{r}
#factor to reorder plot
metadata_with_hill$Mangrove.system <- factor(metadata_with_hill$Mangrove.system)
metadata_with_hill$Depths  <- 
  factor(metadata_with_hill$Depths)

```

```{r}
# Change fossil lagoon reference
metadata_with_hill$Mangrove.system <- relevel(metadata_with_hill$Mangrove.system, ref = "Fossil Lagoon")
```

### q0 Observed richness

```{r}
# lm
q0lm <- lm(q0 ~ log(Effective_reads) + Mangrove.system + Depths, data = metadata_with_hill)

summary(q0lm)

```

```{r}
# check model reliability
shapiro.test(residuals(q0lm))
```

```{r}
#get differences with emmeans
q0_lm_means <- emmeans(q0lm, pairwise ~ Mangrove.system)

# get significance letters
#library(multcomp)
cld_results <- cld(object = q0_lm_means$emmeans, Letters = letters)

# Convert to data frame
q0_emmeans_df <- as.data.frame(cld_results)
q0_emmeans_df

# Reorder samples
metadata_with_hill$Mangrove.system <- factor(
  metadata_with_hill$Mangrove.system,
  levels = c("Fossil Lagoon", "San Pedro River", 
             "Términos Lagoon"))
```

```{r}
# boxplot with significance
q0plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system, y = q0, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q0_emmeans_df, aes(y = emmean, label = .group), vjust = -3,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q0 = Observed", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none", axis.text.x = element_blank())

q0plot
```

### q1 Diversity Exp Shannon

```{r}
# linear model
q1lm <- lm(q1 ~ log(Effective_reads) + Mangrove.system + Depths, data = metadata_with_hill)

# summary
summary(q1lm)

#check model reliability
shapiro.test(residuals(q1lm))

# significance
q1_lm_means <- emmeans(q1lm, pairwise ~ Mangrove.system)
cld_results <- cld(object = q1_lm_means$emmeans, Letters = letters)

# Convert to data frame
q1_emmeans_df <- as.data.frame(cld_results)
q1_emmeans_df

```

```{r}
# boxplot
q1plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system , y = q1, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q1_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +   labs(title = "q1 = Exp Shannon", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_blank())

q1plot
```

### q2 Dominance Inv Simpson

```{r}
# Model
q2lm <- lm(q2 ~ log(Effective_reads) + Mangrove.system + Depths, data = metadata_with_hill)

# summary
summary(q2lm)

#check model reliability
shapiro.test(residuals(q2lm))

# get significance
q2_lm_means <- emmeans(q2lm, pairwise ~ Mangrove.system)
cld_results <- cld(object = q2_lm_means$emmeans, Letters = letters)

# Convert to data frame
q2_emmeans_df <- as.data.frame(cld_results)
q2_emmeans_df
```

```{r}
# boxplot
q2plot <- ggplot(metadata_with_hill, aes(x = Mangrove.system, y = q2, fill = Mangrove.system)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q2_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
            color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q2 = Inv Simpson", y = NULL, x = NULL,
  caption = "lm (Depth sequencing, Mangrove system and Depths") +
  theme_classic() + scale_fill_manual(values = loc_colors) +
  theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_text(size = 9),
    plot.caption = element_text(size = 7, face = "italic"))

q2plot
```

### Combine plot

```{r}
#Combine alpha diversity plots

ytitle <- ggdraw() + draw_label("Effective number of ASVs",
                                angle = 90, size = 14 )
q012plot <- plot_grid(q0plot, q1plot, q2plot,
                      labels = c("A)", "", ""), ncol = 1)
q012y_plot <- plot_grid(ytitle, q012plot, ncol = 2, rel_widths = c(0.05, 1))

# show
q012y_plot

# save
ggsave("Figures/Depth_comparison/alphadiv_per_MS.pdf",
       q012y_plot, width = 6.3, height = 9.9)
```

### Save rds for alpha diversity plot

```{r}
saveRDS(q012y_plot, "rds/Depth_comparison/alpha_diversity_MS_plot.rds")
```

## 06.Get significant differences for Depths

```{r}
#factor to reorder plot
metadata_with_hill$Mangrove.system <- factor(metadata_with_hill$Mangrove.system)
metadata_with_hill$Depths  <- 
  factor(metadata_with_hill$Depths)

# Change fossil lagoon reference
metadata_with_hill$Depths <- relevel(metadata_with_hill$Depths, ref = "Surface")
```

### q0 Observed richness

```{r}
# lm
q0lm <- lm(q0 ~ log(Effective_reads) + Depths + Mangrove.system, data = metadata_with_hill)

summary(q0lm)
```

```{r}
# check model reliability
shapiro.test(residuals(q0lm))
```

```{r}
#get differences with emmeans
q0_lm_means <- emmeans(q0lm, pairwise ~ Depths)

# get significance letters

cld_results <- cld(object = q0_lm_means$emmeans, Letters = letters)

# Convert to data frame
q0_emmeans_df <- as.data.frame(cld_results)
q0_emmeans_df

# Reorder samples
metadata_with_hill$Depths <- factor(
  metadata_with_hill$Depths,
  levels = c("Surface", "mid", 
             "Deep"))
```

```{r}
# boxplot with significance
q0plot <- ggplot(metadata_with_hill, aes(x = Depths, y = q0, fill = Depths)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q0_emmeans_df, aes(y = emmean, label = .group), vjust = -3,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q0 = Observed", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = depth_colors) +
  theme(legend.position = "none", axis.text.x = element_blank())

q0plot
```

### q1 Diversity Exp Shannon

```{r}
# linear model
q1lm <- lm(q1 ~ log(Effective_reads) + Depths + Mangrove.system, data = metadata_with_hill)

# summary
summary(q1lm)

#check model reliability
shapiro.test(residuals(q1lm))

# significance
q1_lm_means <- emmeans(q1lm, pairwise ~ Depths)
cld_results <- cld(object = q1_lm_means$emmeans, Letters = letters)

# Convert to data frame
q1_emmeans_df <- as.data.frame(cld_results)
q1_emmeans_df

```

```{r}
# boxplot
q1plot <- ggplot(metadata_with_hill, aes(x = Depths , y = q1, fill = Depths)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q1_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
          color = "black", fontface = "bold", position = position_dodge(0.9)) +   labs(title = "q1 = Exp Shannon", y = NULL, x = NULL) + #,
  #caption = "letters by lm emmeans") +
  theme_classic() + scale_fill_manual(values = depth_colors) +
  theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_blank())

q1plot
```

### q2 Dominance Inv Simpson

```{r}
# Model
q2lm <- lm(q2 ~ log(Effective_reads) + Depths + Mangrove.system, data = metadata_with_hill)

# summary
summary(q2lm)

#check model reliability
shapiro.test(residuals(q2lm))

# get significance
q2_lm_means <- emmeans(q2lm, pairwise ~ Depths)
cld_results <- cld(object = q2_lm_means$emmeans, Letters = letters)

# Convert to data frame
q2_emmeans_df <- as.data.frame(cld_results)
q2_emmeans_df
```

```{r}

# boxplot
q2plot <- ggplot(metadata_with_hill, aes(x = Depths, y = q2, fill = Depths)) +
  geom_boxplot() +
  #geom_jitter(aes(fill = Mangrove.system), width = 0.1, alpha = 0.4) +
  geom_text(data = q2_emmeans_df, aes(y = emmean, label = .group), vjust = -5,
            color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "q2 = Inv Simpson", y = NULL, x = NULL,
  caption = "lm (Depth sequencing, Dephts and Mangrove system") +
  theme_classic() + scale_fill_manual(values = depth_colors) +
  theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_text(size = 9),
    plot.caption = element_text(size = 7, face = "italic"))

q2plot
```

### Combine plot

```{r}
#Combine alpha diversity plots

ytitle <- ggdraw() + draw_label("Effective number of ASVs",
                                angle = 90, size = 14 )
q012plot <- plot_grid(q0plot, q1plot, q2plot,
                      labels = c("A)", "", ""), ncol = 1)
q012y_plot <- plot_grid(ytitle, q012plot, ncol = 2, rel_widths = c(0.05, 1))

# show
q012y_plot

# save
ggsave("Figures/Depth_comparison/alphadiv_per_Depths.pdf",
       q012y_plot, width = 6.3, height = 9.9)
```

### Save rds for Depths alpha diversity plot

```{r}
saveRDS(q012y_plot, "rds/Depth_comparison/alpha_diversity_Depths_plot.rds")
```
