---
title: "Beta Diversity"
---

::: callout-note
For beta diversity analysis, we conducted NMDS and PERMANOVA tests to observe the influence of Depth parameters in the structure of the microbial community
:::

```{r}
# load libraries
library(vegan)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(pairwiseAdonis)
library(reshape2)
```

```{r}
#load data
physeq_qiime3 <- readRDS("rds/Depth_comparison/physeq_qiime3.rds")
```

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Términos Lagoon" = "#329D9C")
```

```{r}
# Extract data from phyloseq object

otu_data <- otu_table(physeq_qiime3, taxa_are_rows = TRUE)
metadata <- as(sample_data(physeq_qiime3), "data.frame")

sample_data <- data.frame(sample_data(physeq_qiime3))
```

## 01. NMDS

### Get distances and stress

```{r}
set.seed(123)
bray=vegdist(t(otu_data), method = "bray")
nmds_source_bray = vegan::metaMDS(bray, trymax = 20, k = 2)

# stress
var_stress_nmds_bray <- round(nmds_source_bray$stress, 5)

# score
scores_source_bray= nmds_source_bray %>% vegan::scores()
```

### Plot

```{r}
nmds_plot_bray <- ggplot() +
  geom_point(data=data.frame(scores_source_bray) %>%
               rownames_to_column(var = "SampleID")%>%
               left_join(metadata, by = "SampleID"),
             aes(x=NMDS1, y=NMDS2,  color = Mangrove.system, 
                 shape = Depths),
             size=5) + theme_linedraw()+
  scale_fill_manual(values = loc_colors)+
  scale_color_manual(values = loc_colors)+
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(colour = "black", size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right",
        legend.box = "vertical")+ theme_classic() +
  labs(title="Bray-Curtis distances") +
  ylab("NMDS2")+xlab("NMDS1")

## stress legend
nmds_plot_bray <- nmds_plot_bray +
  annotate("text", x = Inf, y = -Inf,
           label = paste("Stress:", var_stress_nmds_bray),
           hjust = 1.1, vjust = -1.1, size = 4)

# show plot
nmds_plot_bray

```

## 02. PERMANOVA

```{r}
# Extract and transpose OTU table
otu_mat <- t(otu_table(physeq_qiime3))

# PERMANOVA with adonis2
set.seed(123)
adonis_result <- adonis2(otu_mat ~ Mangrove.system * Depths,
                         data = sample_data,
                         method = "bray",
                         permutations = 999,
                         by = "terms")

# show adonis result
print(adonis_result)
```

### Plot with PERMANOVA results

```{r}
#add to plot
nmds_plot_bray_perm <- nmds_plot_bray + annotate("text", x = Inf, y = -Inf,
           label = paste("PERMANOVA:",
                         "Mangrove system: R² = 0.328, p = 0.001\n",
                         "Depths: R² = 0.047, p = 0.007\n",
                         "Residual: R² = 0.540\n", 
                         "BetaDisper Mangrove.system: p-value = 0.001\n",
                         "BetaDisper Depths: p-value =  0.728"),
           hjust = 1.1, vjust = -7, size = 3)

# Mostrar el gráfico
print(nmds_plot_bray_perm)

#save
ggsave("Figures/Depth_comparison/beta-diversity.pdf", nmds_plot_bray_perm, width = 7.3, height = 5.5)
```

## 03. Post-hoc Adonis

### Pairwise Adonis

```{r}
pairwise_adonis <- pairwiseAdonis::pairwise.adonis2(otu_mat ~ 
                  Mangrove.system, data = sample_data, 
                  method = "bray",  permutations = 999)

print(pairwise_adonis)
```

### P-values matrix

```{r}
#  p-values of pairwise_adonis
p_values <- c("San Pedro River_vs_Fossil Lagoon" = 0.001,
  "San Pedro River_vs_Términos Lagoon" = 0.001,
  "Fossil Lagoon_vs_Términos Lagoon" = 0.001
)

# matrix of p-values
locations <- c("San Pedro River", "Fossil Lagoon", "Términos Lagoon")
p_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                   dimnames = list(locations, locations))

# Fill matrix with p-values (only for the upper half then make symmetrical)
pairs <- strsplit(names(p_values), "_vs_")
for (i in seq_along(pairs)) {
  row_name <- pairs[[i]][1]
  col_name <- pairs[[i]][2]
  p_matrix[row_name, col_name] <- p_values[i]
  p_matrix[col_name, row_name] <- p_values[i]  # symmetrical
}
diag(p_matrix) <- 1  # compare with itself

# long format
p_melt <- melt(p_matrix)
```

### Heatmap of Adonis Posthoc

```{r}
# heatmap
heatmap_plot <- ggplot(p_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0.025, limits = c(0, 0.05),
                       name = "p-value") +
  geom_text(aes(label = sprintf("%.3f", value)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "Pairwise PERMANOVA post-hoc)",
       x = "Mangrove system", y = "Mangrove system") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14))

# show
print(heatmap_plot)
```

## 04. Beta Disper

```{r}
# Bray distance matrix
dist_matrix <- phyloseq::distance(physeq_qiime3, method = "bray")

```

### By Mangrove System

```{r}
# betadisper
disp_location <- betadisper(dist_matrix, sample_data$Mangrove.system)
print(summary(disp_location))

anova_beta_disp <- anova(disp_location)

# Show dispersion by group
disp_location$group.distances

```

#### Permutest

```{r}
# Significance permutest
perm_location <- permutest(disp_location, permutations = 999)
print(perm_location)
```

```{r}
# Plots
plot(disp_location, main = "Beta Dispersión por Mangrove system")
boxplot(disp_location, main = "Distancias al centroide por Mangrove system")
```

### By Depths

```{r}
# Betadisper Mangrove_type
disp_Depths <- betadisper(dist_matrix, sample_data$Depths)
print(summary(disp_Depths))

anova_beta_disp_depth <- anova(disp_Depths)

# Show dispersion by group
disp_Depths$group.distances

# Significance permutest
perm_Depths <- permutest(disp_Depths, permutations = 999)
print(perm_Depths)
```

```{r}
# plots
plot(disp_Depths, main = "Betadisper  Depths")
boxplot(disp_Depths, main = "Dist to centroid Depths")
```

## 05. Plot with PERMANOVA, pairwise PERMANOVA and BetaDisper

```{r}
beta_text_ms <- paste("Betadisper Mangrove system, p-value:", 
                   round(anova_beta_disp$`Pr(>F)`[1], 3))
beta_text_depth <- paste("Betadisper Depth, p-value:", 
                   round(anova_beta_disp_depth$`Pr(>F)`[1], 3))

# PERMANOVA text
# Results
perm_results <- data.frame(
  Term = c("Location","Depths", "Residual"),
  R2 = c(adonis_result$R2[1], adonis_result$R2[2], adonis_result$R2[3]),
  p_value = c(adonis_result$`Pr(>F)`[1], adonis_result$`Pr(>F)`[2], NA))

# Text
perm_text <- paste(
  "PERMANOVA:",
  paste("Mangrove system: R² =", round(perm_results$R2[1], 3), ", p =", perm_results$p_value[1]),
  paste("Depths: R² =", round(perm_results$R2[2], 3), ", p =", perm_results$p_value[2]), sep = "\n")
```

```{r}
# Add texts to NMDS
nmds <- nmds_plot_bray +
  annotate(
    "text", x = Inf, y = Inf, 
    label = paste(perm_text, beta_text_ms, beta_text_depth, sep = "\n"),
    hjust = 1.1, vjust = 1.2, size = 3, color = "black")

nmds
```

```{r}
ggsave("Figures/Depth_comparison/NMDS.pdf", nmds, bg='transparent', width = 6, height = 4.5)
```

## Save rds for beta diversity plot

```{r}
saveRDS(nmds, "rds/Depth_comparison/beta-diversity-Depth-plot.rds")
```

##  Supplementary

```{r}
library(dplyr)

# Extract PERMANOVA 
tab_perm <- as.data.frame(adonis_result) %>%
  mutate(Factor = rownames(.)) %>%
  filter(!Factor %in% c("Total", "Residual")) %>%
  dplyr::select(Factor, Df, R2, F.Model = F, p_value = `Pr(>F)`)

# Extract BETADISPER 
tab_disper <- data.frame(
  Factor = c("Mangrove.system", "Depths"),
  Dispersion_p = c(anova(disp_location)$`Pr(>F)`[1], 
                   anova(disp_Depths)$`Pr(>F)`[1])
)

# join
tableS6 <- left_join(tab_perm, tab_disper, by = "Factor") %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

# Guardar
write.csv(tableS6, "Tables/Depth_comparison/Supplementary_Table_S6_Beta_Depth_Combined.csv", row.names = FALSE)
```
