---
title: "Data Prepare"
---

::: callout-note
## How are coastal and inland mangroves similar, considering their depth ❓

To answer this question, only samples that met the selection criteria were selected for comparison with each other. These samples were as follows:

1.  They hailed from Terminos Lagoon, Fossil Lagoon and San Pedro River

2.  They hailed from strip, in case of Terminos Lagoon

3.  They hailed from the flooding season, in case of Terminos Lagoon

4.  They hailed from every depth registered
:::

## 01.Select Samples

```{r}
library(ggplot2)
library(qiime2R)
library(phyloseq)
library(microbiome)
#Create phyloseq objects with qiime2 artifacts
physeq_qiime3 <- qza_to_phyloseq(
  features = "data/cluster_table_filter_freq218_emcEstero.qza",
  taxonomy = "data/taxonomyEstero.qza",
  metadata = "data/metadata14.tsv")
```

```{r}
#Filter metadata
# Extract metadata from phyloseq object
metadat <- data.frame(phyloseq::sample_data(physeq_qiime3), check.names = FALSE)
```

```{r}
#Filter metadata by Study_zone column
filtered_metadat <- metadat[metadat$Study_zone %in% c("Estero_pargo", "Rio San Pedro", "Laguna Cacahuate"), ]
```

```{r}
#Filter again by Ecological Type column, selecting Fringe parameter
filtered_metadat2 <- filtered_metadat[filtered_metadat$Ecological_type %in% c("Fringe"), ]
```

```{r}
#Filter again by season column to include all samples that were collected form a flood season
filtered_metadat3 <- filtered_metadat2[filtered_metadat2$season %in%c("flood"),]
```

```{r}
#Combine filtered metadata with phyloseq object to create a filtered phyloseq object
filtered_physeq2 <- prune_samples(rownames(filtered_metadat3), physeq_qiime3)
```

```{r}
#Save your new phyloseq object
saveRDS(filtered_physeq2, file = "rds/Depth_comparison/physeq_objeto_Estero.rds") 
```

## 02.Data Preparation

```{r}
#load
physeq_qiime <- readRDS("rds/Depth_comparison/physeq_objeto_Estero.rds")

```

### 02.1 Remove empty ASV's and singletons

```{r}
#| message: false
#| warning: false
#check data
library(microbiome)
library(dplyr)
```

```{r}
#library(microbiome)
microbiome::summarize_phyloseq(physeq_qiime)
```

```{r}
# get sums
taxa_sums <- taxa_sums(physeq_qiime)
# Identify ASVs with sum 0 present in all samples
zero_taxa <- taxa_names(physeq_qiime)[taxa_sums == 0]
# Remove ASVs with sum 0
physeq_qiime2 <- prune_taxa(!taxa_names(physeq_qiime) %in%
                                zero_taxa, physeq_qiime)

# Remove singletons
physeq_qiime3 <- filter_taxa(physeq_qiime2, function(x) sum(x) > 1, TRUE)
```

## 02.2 Rename Mangrove system

```{r}
# Extract metadata
metadata <- as(sample_data(physeq_qiime3), "data.frame")

# Rename specific levels in the Mangrove system column
#library(dplyr)
metadata <- metadata %>%
  mutate(`Mangrove.system` = case_when(
    Study_zone == "Rio San Pedro" ~ "San Pedro River",
    Study_zone == "Laguna Cacahuate" ~ "Fossil Lagoon",
    Study_zone == "Estero_pargo" ~ "Términos Lagoon",
    TRUE ~ Study_zone))

```

## 02.4 Rename by depth

```{r}
#library(dplyr)
metadata <- metadata %>%
  mutate(`Depths` = case_when(
    depth %in% c("0-15", "5") ~ "Surface",
    depth %in% c("16-30", "20") ~ "mid",
    depth %in% c("31-45", "50-70", "40") ~ "Deep",
  TRUE ~ NA_character_))
```

## 02.3 Create color pallete

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Términos Lagoon" = "#329D9C")
```

## 02.4 Check samples

```{r}

samples_plot <- ggplot(metadata, aes(x = Depths, y = ..count..,
                                     fill = Mangrove.system)) +
  geom_bar(position = "dodge") +
  labs(title = "Samples of the analysis", 
       y = "Number of Samples",
       x = "Mangrove system") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0)) +
  scale_fill_manual(values = loc_colors)

# show plot
print(samples_plot)
```

## 02.5 Save object in RDS for future analysis

```{r}
saveRDS(physeq_qiime3, "rds/Depth_comparison/physeq_qiime3.rds")
```
