---
title: "Alpha Diversity"
---

# Load libraries and prepare data

```{r}
#| label: libraries
#| message: false
#| warning: false
#Load libraries
library(phyloseq)
library(ape)
library(qiime2R)
library(hilldiv)
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(cowplot)
library(vegan)
library(emmeans)
library(multcomp)
library(lme4)
library(lmerTest)
library(easystats)
```

```{r}
#load data
ps <- readRDS("rds/interior_impaired/physeq.rds")
```

```{r}
# Extract data from phyloseq object
#library(phyloseq)
otu_table <- otu_table(ps, taxa_are_rows = TRUE)
metadata <- as(sample_data(ps), "data.frame")
```

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Impaired Términos Lagoon" = "gray40")

#Mangrove type colors
mt_colors <- c("Interior"= "#A7fcc1",
                "Coastal" = "#329D9C")

#Depth colors
depth_colors <- c("Surface" = "#f5e5c4",
                 "Mid" = "#baa179",
                 "Deep" = "#80673b")
```

# 01. Get Hill numbers

```{r}
# library(hilldiv)
# library(tidyverse)

# Get hill numbers
q0 <- hill_div(otu_table, qvalue = 0) 
q1 <- hill_div(otu_table, qvalue = 1) 
q2 <- hill_div(otu_table, qvalue = 2)
```

```{r}
# Merge metadata with Hill numbers
q012_all <- cbind(q0, q1, q2) %>% as.data.frame() %>% rownames_to_column(var = "SampleID")
metadata_with_hill <- q012_all %>%          
  inner_join(metadata, by = c("SampleID"="SampleID"))

#save table
write.table(metadata_with_hill, "Tables/interior_impaired/Metadata_with_hill.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)

# show
metadata_with_hill %>% head()
```

# 02. Get Hill means

```{r}
# Reorder q values  
meta_qs <- metadata_with_hill %>%
    pivot_longer(cols = q0:q2, names_to = "q", values_to = "value") %>%
    filter(q %in% c("q0", "q1", "q2")) %>%
    mutate(
        qs = case_when(
            q == "q0" ~ "q0=Observed",
            q == "q1" ~ "q1=Exp Shannon",
            q == "q2" ~ "q2=Inv Simpson"))

#Get means of Hill numbers
means <- meta_qs %>% group_by(Depths, Mangrove_system, Mangrove_type, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')

#save table
write.table(means, "Tables/interior_impaired/Hill_means_sd.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)

print(means)
```

```{r}
#Get means of Hill numbers
hillmeansloc <- meta_qs %>% group_by(Depths, Mangrove_system, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')
hillmeansloc

#save table
write.table(hillmeansloc, "Tables/interior_impaired/Hill_means_sd_Mangrove_system.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)
```

```{r}
#Get means of Hill numbers
hillmeanstype <- meta_qs %>% group_by(Depths, Mangrove_type, qs) %>%
  summarise(means = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = 'drop')
hillmeanstype

#save table
write.table(hillmeanstype, "Tables/interior_impaired/Hill_means_sd_type.tsv", quote = FALSE, sep = "\t", row.names = TRUE, col.names=TRUE)
```

# 03. Get sequencing effort

```{r}
#| message: false
#| warning: false
# faster
#library(vegan)
mat <- as(t(otu_table(ps)), "matrix")
raremax <- min(rowSums(mat))

# loc colors
Mangrove_system_factor <- sample_data(ps)$Mangrove_system
sample_colors <- loc_colors[as.character(Mangrove_system_factor)] 

# font size
par(cex = 0.6) 

# Rarefaction curves
system.time(rarecurve(mat, step = 100, sample = raremax, 
                      col = sample_colors, label = FALSE))

# save
pdf("Figures/interior_impaired/Sample_effort.pdf")
system.time(rarecurve(mat, step = 100, sample = raremax, 
                      col = sample_colors, label = FALSE))
dev.off()

#slow but pretty
#library(ranacapa)
#library(ggplot2)

#effort_sampling<- ggrare(ps3, step = 100, color = "sample", label = "sampleID", se = TRUE)
#rarecurves_plot <- effort_sampling + facet_wrap(~sample)
```

# 04. Explore hill distribution

::: callout-tip
## Important

Following the calculation of the diversity indices and confirmation of adequate sequencing effort, an investigation was conducted on their distribution along the sampled sites and depths.
:::

```{r}
# factor to reorder plot
metadata_with_hill$Mangrove_system <- factor(metadata_with_hill$Mangrove_system, levels = c("Fossil Lagoon", "San Pedro River", "Impaired Términos Lagoon"))
metadata_with_hill$Depth  <- 
  factor(metadata_with_hill$Depths, 
         levels = c("Surface", "Mid", "Deep"))


#plot
hill_barplot_explore <- metadata_with_hill %>%
  pivot_longer(cols = q0:q2, names_to = "q", values_to = "value") %>%
 filter(q %in% c("q0", "q1", "q2")) %>%
    mutate(
        qs = case_when(
            q == "q0" ~ "q0=Observed",
            q == "q1" ~ "q1=Exp Shannon",
            q == "q2" ~ "q2=Inv Simpson")) %>%
  ggbarplot(
    x = "Mangrove_system",
    y = "value",
    add = "mean_se",
    facet.by = c("qs", "Depth"),
    fill = "Mangrove_system") +
  scale_fill_manual(values = loc_colors) +
  geom_jitter(size = 0.8, 
              position = position_jitter(width = 0.1)) +
  facet_grid(rows = vars(qs), cols = vars(Depth), 
             scales = "free_y") +
  theme(
    strip.background =  element_blank(),
    strip.text.x = element_text(face= "bold", size = 13,
                margin = margin(0.5, 0, 0.5, 0)),
    strip.text.y = element_text(face= "bold", size = 14,
                angle = 0, margin = margin(0, 0.5, 0, 0.5)),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(colour = "black"),
    #panel.border = element_blank(),
    panel.spacing.x = unit(0.5, "lines"),
    panel.background = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(colour = "black", size = 11),
    legend.position = "bottom") +
  labs(
    fill = "Mangrove_system",
    y = "",
    x = "",
    title = "Biodiversity across Sites and Depths") + 
   theme(legend.text = element_text(size = 11))

hill_barplot_explore

ggsave("Figures/interior_impaired/Hill_diversity_across_Sites_and_Depths.pdf", hill_barplot_explore, width = 11, height = 11)
```

::: callout-tip
At this juncture, the available data permits only the observation of the distribution of the phenomenon under study. However, it must be noted that comparisons at this stage may prove misleading due to the presence of other variables that must be considered to accurately assess the observed differences.
:::

# 05. Check alpha diversity correlation with sequencing depth

```{r}
#Get effective reads
Effective_reads <- sample_sums(otu_table)
metadata_with_hill$Effective_reads <- Effective_reads
```

**Alpha diversity depth correlation to order q=0**

```{r}
# library(ggpubr)

q0_vs_depth <- ggscatter(metadata_with_hill,
                x = "Effective_reads", y = "q0", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=0", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient. 
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")
   )+
  theme(legend.title = element_blank(), legend.position = "none")
```

**Alpha diversity depth correlation to order q=1**

```{r}
q1_vs_depth <- ggscatter(metadata_with_hill, 
                x = "Effective_reads", y = "q1", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=1", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")) +
  theme(legend.title = element_blank(), 
        legend.position = "none")

```

**Alpha diversity depth correlation to order q=2**

```{r}
q2_vs_depth <- ggscatter(metadata_with_hill, 
                x = "Effective_reads", y = "q2", 
                xlab= "Sequencing depth (number of reads)",
                ylab = "q=2", add = "reg.line",
   add.params = list(color = "#114e9d", fill = "lightgray"),
   conf.int = TRUE, # Add confidence interval
   cor.coef = TRUE, # Add correlation coefficient
   cor.coeff.args = list(method = "pearson", label.x = 3, label.sep = "\n")) +
  theme(legend.title = element_blank(), 
        legend.position = "none")
```

```{r}
# library(ggplot2)
# library(cowplot)

#Combine plot
title_corr_plot <- ggdraw() + 
  draw_label("Alpha diversity depth correlation")

correlation_plot_q012 <- plot_grid(title_corr_plot,
                            q0_vs_depth, 
                            q1_vs_depth, 
                            q2_vs_depth,
                            ncol = 1, 
                            rel_heights = c(0.15, 1, 1, 1))

correlation_plot_q012

# #save plot
ggsave("Figures/interior_impaired/alpha_depth_correlation.pdf", correlation_plot_q012, width = 8, height = 8)
```

::: callout-caution
## Take note

We found a positive correlation between *Observed richness* and *Shannon Diversity* with sequencing depth, indicating that these variables should be considered when conducting meaningful comparisons of this index.
:::

------------------------------------------------------------------------

::: callout-caution
The sequencing depth, presence of repeated measures, and imbalances in the project data necessitates model-based comparison. In this instance, the utilization of linear mixed models is a good way to evaluate effects and differences.
:::

# 06. Get significant differences

::: callout-caution
## Remember

The significance of the observed differences among mangrove systems was determined by employing a linear mixed model, given the imbalanced sample numbers.
:::

------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
# Change fossil lagoon reference
metadata_with_hill$Mangrove_system <- relevel(metadata_with_hill$Mangrove_system, ref = "Impaired Términos Lagoon")
```

## By depth

### q0 Observed Richness

```{r}
# lmm
q0lmer <- lmer(q0 ~ log(Effective_reads) + Depth + Mangrove_system + (1 | ID), data = metadata_with_hill)
# summary
summary(q0lmer)
```

```{r}
# likelihood-ratio test
lrt_random_q0 <- rand(q0lmer, nsim = 1000)
lrt_random_q0

#effects
anova(q0lmer)
```

```{r}
#| message: false
#| warning: false
# get differences with emmeans
q0_lmer_means <- emmeans(q0lmer, pairwise ~ Depth)

# get significance letters
cld_results <- cld(object = q0_lmer_means$emmeans, Letters = letters)

# Convert to data frame
q0_emmeans_df <- as.data.frame(cld_results)
q0_emmeans_df
```

```{r}
# boxplot with significance
q0plot <- ggplot(metadata_with_hill, aes(x = Depth, y = q0, fill = Depth)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Depth), width = 0.1, alpha = 0.4) +
  geom_text(data = q0_emmeans_df, aes(y = emmean, label = .group), vjust = -5, color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "Observed Richness", y = NULL, x = NULL,
       caption = "lmer (depth sequencing, layers, location and sediment core)") +
    theme_classic() + scale_fill_manual(values = depth_colors) +
    theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_text(size = 9),
          plot.caption = element_text(size = 7, face = "italic"))

q0plot
```

### q1 Diversity Exp Shannon

```{r}
# Fitted linear mixed model
q1lmer <- lmer(q1 ~ log(Effective_reads) + Depth + Mangrove_system  + (1 | ID), data = metadata_with_hill)
# summary
summary(q1lmer)
```

```{r}
# lmerTest
lrt_random_q1 <- rand(q1lmer, nsim = 1000)

# Summary
lrt_random_q1

# Efects
anova(q1lmer)
```

```{r}
# Sig dif emmeans
q1_lmer_means <- emmeans(q1lmer, pairwise ~ Depth)

# letters
cld_resultsq1 <- cld(object = q1_lmer_means$emmeans, Letters = letters)

# Convert df
emmeans_dfq1 <- as.data.frame(cld_resultsq1)
emmeans_dfq1
```

```{r}
# plot
q1plot <- ggplot(metadata_with_hill, aes(x = Depth, y = q1, fill = Depth)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Depth), width = 0.1, alpha = 0.4) +
  geom_text(data = emmeans_dfq1, aes(y = emmean, label = .group), vjust = -9, color = "black", fontface = "bold",position = position_dodge(0.9)) +
  labs(title = "Exp Shannon", y = NULL, x = NULL ) + #,
       #caption = "letters by lmer emmeans") +
    theme_bw() + scale_fill_manual(values = depth_colors) +
    theme(legend.position = "none")
q1plot
```

### q2 Dominance Inv Simpson

```{r}
# Fitted linear mixed model
q2lmer <- lmer(q2 ~ Depth + Mangrove_system + (1 | ID), data = metadata_with_hill)
#summary
summary(q2lmer)
```

```{r}
# lmerTest
lrt_random_q2 <- rand(q2lmer, nsim = 1000)

# Summary
lrt_random_q2 

# Efects
anova(q2lmer)
```

```{r}
# Sig dif emmeans
q2_lmer_means <- emmeans(q2lmer, pairwise ~ Depth)

# letters
cld_resultsq2 <- cld(object = q2_lmer_means$emmeans, Letters = letters)

# Convert df
emmeans_dfq2 <- as.data.frame(cld_resultsq2)
emmeans_dfq2
```

```{r}
# plot
q2plot <- ggplot(metadata_with_hill, aes(x = Depth, y = q2, fill = Depth)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Depth), width = 0.1, alpha = 0.4) +
  geom_text(data = emmeans_dfq2, aes(y = emmean, label = .group), vjust = -9, color = "black", fontface = "bold",position = position_dodge(0.9)) +
  labs(title = "q2 = Inv Simpson", y = NULL, x = NULL ) + #,
       #caption = "letters by lmer emmeans") +
    theme_bw() + scale_fill_manual(values = depth_colors) +
    theme(legend.position = "none")
q2plot
```

## By Mangrove_system

### q0 Observed Richness

```{r}
# lmm
q0lmerloc <- lmer(q0 ~ log(Effective_reads) + Mangrove_system + Depth + (1 | ID), data = metadata_with_hill)
# summary
summary(q0lmerloc)
```

```{r}
# Permutation test 
lrt_q0loc <- rand(q0lmerloc, nsim = 1000)
lrt_q0loc

#effects
anova(q0lmerloc)
```

```{r}
# emmeans
q0_lmer_means_loc <- emmeans(q0lmerloc, pairwise ~ Mangrove_system)

# significance letters
q0cld_results_loc <- cld(object = q0_lmer_means_loc$emmeans, Letters = letters)

# Convert to data frame
q0emmeans_df_loc <- as.data.frame(q0cld_results_loc)
q0emmeans_df_loc
```

```{r}
# boxplot with significance
q0plotloc <- ggplot(metadata_with_hill, aes(x = Mangrove_system, y = q0, fill = Mangrove_system)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Mangrove_system), width = 0.1, alpha = 0.4) +
  geom_text(data = q0emmeans_df_loc, aes(y = emmean, label = .group), vjust = -3, color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "Observed Richness", y = NULL, x = NULL ) + #,
       #caption = "letters by lmer emmeans") +
    theme_classic() + scale_fill_manual(values = loc_colors) +
    theme(legend.position = "none", axis.text.x = element_blank())

q0plotloc
```

### q1 Exp Shannon diversity

```{r}
#| label: q1lmerloc
# lmm
q1lmerloc <- lmer(q1 ~ log(Effective_reads) + Mangrove_system + Depth + (1 | ID), data = metadata_with_hill)
# summary
summary(q1lmerloc)
```

```{r}
# LRT test 
lrt_q1loc <- rand(q1lmerloc, nsim = 1000)
lrt_q1loc

#effects
anova(q1lmerloc)
```

```{r}
# Differences emmeans
q1lmer_means_loc <- emmeans(q1lmerloc, pairwise ~ Mangrove_system)

# significance letters
q1cld_results_loc <- cld(object = q1lmer_means_loc$emmeans, Letters = letters)

# Convert to data frame
q1emmeans_df_loc <- as.data.frame(q1cld_results_loc)
q1emmeans_df_loc
```

```{r}
# boxplot with significance
q1plotloc <- ggplot(metadata_with_hill, aes(x = Mangrove_system, y = q1, fill = Mangrove_system)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Mangrove_system), width = 0.1, alpha = 0.4) +
  geom_text(data = q1emmeans_df_loc, aes(y = emmean, label = .group), vjust = -3, color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "Exp Shannon", y = NULL, x = NULL,
       caption = "lmer (Depth sequencing, Mangrove system, layers and sediment core)") +
    theme_classic() + scale_fill_manual(values = loc_colors) +
    theme(legend.position = "none",
          axis.text.y = element_text(size = 9),
          axis.text.x = element_text(size = 9),
    plot.caption = element_text(size = 7, face = "italic"))

q1plotloc
```

### q2 Dominance Inv Simpson

```{r}
# lmm
q2lmerloc <- lmer(q2 ~ Mangrove_system + Depth + (1 | ID), data = metadata_with_hill)
# summary
summary(q2lmerloc)
```

```{r}
# LRT test
lrt_q2loc <- rand(q2lmerloc, nsim = 1000)
lrt_q2loc

#effects
anova(q2lmerloc)
```

```{r}
# Differences  emmeans
q2lmer_means_loc <- emmeans(q2lmerloc, pairwise ~ Mangrove_system)

# Significance 
q2cld_results_loc <- cld(object = q2lmer_means_loc$emmeans, Letters = letters)

# Convertir a data frame
q2emmeans_df_loc <- as.data.frame(q2cld_results_loc)
q2emmeans_df_loc
```

```{r}
# boxplot with significance
q2plotloc <- ggplot(metadata_with_hill, aes(x = Mangrove_system, y = q2, fill = Mangrove_system)) +
  geom_boxplot() +
  geom_jitter(aes(fill = Mangrove_system), width = 0.1, alpha = 0.4) +
  geom_text(data = q2emmeans_df_loc, aes(y = emmean, label = .group), vjust = -3, color = "black", fontface = "bold", position = position_dodge(0.9)) +
  labs(title = "Inv Simpson", y = NULL, x = NULL ) + #,
       #caption = "letters by lmer emmeans") +
    theme_classic() + scale_fill_manual(values = loc_colors) +
    theme(legend.position = "none")

q2plotloc
```

## Combine plots

::: callout-caution
## Take note

A combined plot of the Observed richness by depth and Mangrove system, and Shannon diversity by Mangrove system was created because the former indices showed significant differences
:::

```{r}
#Combine plot

title <- ggdraw() + draw_label("Effective number of ASVs", angle = 90, size = 12)
alphaplots_loc <- plot_grid(q0plotloc,
                      ncol = 1)

alphaplots_loc <- plot_grid(title, alphaplots_loc, ncol = 2, rel_widths = c(0.05, 1))    

alphaplots_loc

alphaplots_depth <- plot_grid(title, q0plot, ncol = 2, rel_widths = c(0.05, 1))    

alphaplots_depth
```

::: callout-important
A comparison of alpha diversity indices revealed no significant differences among the mangrove sites. Consequently, incorporation of these into the final plot was unnecessary.
:::

# Save rds plots

```{r}
saveRDS(alphaplots_depth, "rds/interior_impaired/alpha_diversity_depth_plot.rds")

saveRDS(alphaplots_loc, "rds/interior_impaired/alpha_diversity_loc_plot.rds")
```

## Supplementary

### Get depth differences by system

q0

```{r}
# factor
metadata_with_hill$Depths <- factor(metadata_with_hill$Depths, levels = c("Surface", "Mid", "Deep"))

#plot q0
q0_depth_system <- ggplot(metadata_with_hill, aes(x = Depths, y = q0, fill = Depths)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~Mangrove_system, scales = "free_y") + 
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  theme_classic() + 
  scale_fill_manual(values = depth_colors) + 
  labs(title = NULL,
       x = NULL, y = "Observed Richness (q0)") +
  theme(legend.position = "none")
```

**q1**

```{r}
#plot
q1_depth_system <- ggplot(metadata_with_hill, aes(x = Depths, y = q1, fill = Depths)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~Mangrove_system, scales = "free_y") + 
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  theme_classic() + 
  scale_fill_manual(values = depth_colors) + 
  labs(title = NULL,
       x = NULL, y = "Exp Shannon Diversity (q1)") +
  theme(legend.position = "none")
```

**q2**

```{r}
q2_depth_system <- ggplot(metadata_with_hill, aes(x = Depths, y = q2, fill = Depths)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.9) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~Mangrove_system, scales = "free_y") + 
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  theme_classic() + 
  scale_fill_manual(values = depth_colors) + 
  labs(title = NULL,
       x = "Sediment Depth", y = "Inv Simpson Dominance (q2)") +
  theme(legend.position = "none")
```

```{r}
#| message: false
#| warning: false
#Combine alpha diversity plots

q012_depth_system_plot <- plot_grid(q0_depth_system,
                                    q1_depth_system,
                                    q2_depth_system, 
                                    ncol = 1)


# show
q012_depth_system_plot

# svae rds
saveRDS(q012_depth_system_plot, "rds/interior_impaired/alpha_diversity_Depths_by_system_plot.rds")

# save
ggsave("Figures/interior_impaired/Supplementary_alphadiv_per_Depths_and_system.pdf",
       q012_depth_system_plot, width = 6.3, height = 9.9)
```

### Tables

```{r}
library(dplyr)
library(broom.mixed)
library(emmeans)

# ---- Linear Mixed Models ----
tableS7_imp_models <- bind_rows(
  tidy(q0lmerloc) %>% mutate(Metric = "q0"),
  tidy(q1lmerloc) %>% mutate(Metric = "q1"),
  tidy(q2lmerloc) %>% mutate(Metric = "q2")
) %>%
  filter(effect == "fixed") %>%
  dplyr::select(Metric, Term = term, Estimate = estimate,
                Std.Error = std.error, Statistic = statistic,
                p_value = p.value) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

# ---- Post hoc: CLD (medias marginales + letras) ----
get_cld <- function(x, metric, factor_name) {
  d <- as.data.frame(x)
  names(d)[1] <- "Level"
  d$Metric <- metric
  d$Factor <- factor_name
  d
}

tableS7_imp_posthoc <- bind_rows(
  get_cld(q0cld_results_loc, "q0", "Mangrove_system"),
  get_cld(q1cld_results_loc, "q1", "Mangrove_system"),
  get_cld(q2cld_results_loc, "q2", "Mangrove_system"),
  get_cld(cld_results,       "q0", "Depth"),
  get_cld(cld_resultsq1,     "q1", "Depth"),
  get_cld(cld_resultsq2,     "q2", "Depth")
) %>%
  dplyr::select(Metric, Factor, Level, emmean, SE, df,
                lower.CL, upper.CL, Significance = .group) %>%
  mutate(across(where(is.numeric), ~ round(., 3)),
         Significance = trimws(Significance))

rownames(tableS7_imp_posthoc) <- NULL

# ---- Post hoc: contrastes pareados (Tukey) ----
get_contrasts <- function(x, metric, factor_name) {
  as.data.frame(x$contrasts) %>%
    mutate(Metric = metric, Factor = factor_name)
}

tableS7_imp_contrasts <- bind_rows(
  get_contrasts(q0_lmer_means_loc, "q0", "Mangrove_system"),
  get_contrasts(q1lmer_means_loc,  "q1", "Mangrove_system"),
  get_contrasts(q2lmer_means_loc,  "q2", "Mangrove_system"),
  get_contrasts(q0_lmer_means,     "q0", "Depth"),
  get_contrasts(q1_lmer_means,     "q1", "Depth"),
  get_contrasts(q2_lmer_means,     "q2", "Depth")
) %>%
  dplyr::select(Metric, Factor, Contrast = contrast, Estimate = estimate,
                SE, df, t.ratio, p.value) %>%
  mutate(across(c(Estimate, SE, df, t.ratio), ~ round(., 3)),
         p.value = format.pval(p.value, digits = 3, eps = 1e-04))

rownames(tableS7_imp_contrasts) <- NULL

# ---- show ----
tableS7_imp_models
tableS7_imp_posthoc
tableS7_imp_contrasts

# ---- save ----
write.csv(tableS7_imp_models,
          "Tables/interior_impaired/Supplementary_Table_S7_Impairment_LMM.csv",
          row.names = FALSE)
write.csv(tableS7_imp_posthoc,
          "Tables/interior_impaired/Supplementary_Table_S7_Impairment_PostHoc.csv",
          row.names = FALSE)
write.csv(tableS7_imp_contrasts,
          "Tables/interior_impaired/Supplementary_Table_S7_Impairment_Contrasts.csv",
          row.names = FALSE)
```
