---
title: "Beta diversity"
---

::: callout-tip
The analysis of beta diversity was conducted using both NMDS and PERMANOVA to identify the factors that influenced the structure of the microbial community.
:::

------------------------------------------------------------------------

## Load libraries and prepare data

```{r}
#| label: libraries
#| message: false
#| warning: false
# load libraries
library(vegan)
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(pairwiseAdonis)
library(cowplot)
library(reshape2)
```

```{r}
#load data
ps <- readRDS("rds/interior_impaired/physeq.rds")
```

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Impaired Términos Lagoon" = "gray40")

#Depth colors
depth_colors <- c("Surface" = "#f5e5c4",
                 "Mid" = "#baa179",
                 "Deep" = "#80673b")
```

```{r}
# Extract data from phyloseq object
#library(phyloseq)
otu_data <- otu_table(ps, taxa_are_rows = TRUE)
metadata <- as(sample_data(ps), "data.frame")

sample_data <- data.frame(sample_data(ps))
```

## 01. NMDS

### Get distances and stress

::: callout-note
Bray-Curtis dissimilarity distances are a measure of the dissimilarity between microbial communities based on their species composition.
:::

```{r}
#| message: false
#| warning: false
# distances
set.seed(123)
bray=vegdist(t(otu_data), method = "bray")
nmds_source_bray = vegan::metaMDS(bray, trymax = 50, k = 3) 

# stress
var_stress_nmds_bray <- round(nmds_source_bray$stress, 5)

# score
scores_source_bray= nmds_source_bray %>% vegan::scores()
```

### Plot

```{r}
# Reorder Mangrove_system
sample_data(ps)$Mangrove_system <- factor(sample_data(ps)$Mangrove_system, levels = c("Fossil Lagoon", "San Pedro River", "Impaired Términos Lagoon"))
```

```{r}
#| message: false
#| warning: false
nmds_plot_bray <- ggplot() +
  geom_point(data=data.frame(scores_source_bray) %>%
               rownames_to_column(var = "SampleID")%>%
               left_join(sample_data(ps), by = "SampleID"),
             aes(x=NMDS1, y=NMDS2,  color = Mangrove_system, 
                 shape = Depths),
             size=5) + theme_linedraw()+
  scale_fill_manual(values = loc_colors)+
  scale_color_manual(values = loc_colors)+
  theme(axis.text = element_text(colour = "black", size = 12),
        axis.title = element_text(colour = "black", size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.position = "right",
        legend.box = "vertical")+ theme_classic() +
  labs(title="Bray-Curtis distances") +
  ylab("NMDS2")+xlab("NMDS1")

## stress legend
nmds_plot_bray <- nmds_plot_bray +
  annotate("text", x = Inf, y = -Inf,
           label = paste("Stress:", var_stress_nmds_bray),
           hjust = 1.1, vjust = -1.1, size = 3)

# show plot
nmds_plot_bray
```

## 02. PERMANOVA

::: callout-note
To test the effects of Mangrove type and Mangrove system on the Bray-Curtis dissimilarity, a PERMANOVA analysis was conducted using adonis2.
:::

```{r}
# Extract and transpose OTU table
otu_mat <- t(otu_table(ps))

# PERMANOVA with adonis2
set.seed(123)
adonis_result <- adonis2(
  otu_mat ~ Mangrove_system + Depths * Mangrove_type,
  data = sample_data,
  method = "bray",
  permutations = 999,
  strata = metadata$ID,
  by = "terms")

# show adonis result
print(adonis_result)
```

```{r}
# PERMANOVA text
# Results
perm_results <- data.frame(
  Term = c("Location","Depths", "Residual"),
  R2 = c(adonis_result$R2[1], adonis_result$R2[2], adonis_result$R2[3]),
  p_value = c(adonis_result$`Pr(>F)`[1], adonis_result$`Pr(>F)`[2], NA))

# Text
perm_text <- paste(
  "PERMANOVA:",
  paste("Mangrove system: R² =", round(perm_results$R2[1], 3), ", p =", perm_results$p_value[1]),
  paste("Depths: R² =", round(perm_results$R2[2], 3), ", p =", perm_results$p_value[2]), sep = "\n")
```

## 03. Post-hoc adonis

::: callout-note
To test differences between the Mangrove systems a Post-hoc pairwise PERMANOVA was conducted using pairwise.adonis
:::

### Pairwise adonis

```{r}
## Post-hoc Mangrove.system
#library(pairwiseAdonis)
pairwise_adonis <- pairwiseAdonis::pairwise.adonis2(otu_mat ~ 
                  Mangrove_system, data = sample_data, 
                  method = "bray",  permutations = 999)

print(pairwise_adonis)
```

### Extract p-values

```{r}
# Define abbrevs
location_abbrevs <- c(
  "San Pedro River" = "SPR",
  "Fossil Lagoon" = "FL",
  "Impaired Términos Lagoon" = "ITL")

# Extract values
locations <- unique(metadata$Mangrove_system)
p_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                   dimnames = list(locations, locations))
r2_matrix <- matrix(NA, nrow = length(locations), ncol = length(locations),
                    dimnames = list(locations, locations))

for (comp in names(pairwise_adonis)[-1]) {  # Exclude 'parent_call'
  pair <- strsplit(comp, "_vs_")[[1]]
  loc1 <- pair[1]
  loc2 <- pair[2]
  p_matrix[loc1, loc2] <- pairwise_adonis[[comp]]$`Pr(>F)`[1]
  p_matrix[loc2, loc1] <- pairwise_adonis[[comp]]$`Pr(>F)`[1]  # Simetric
  r2_matrix[loc1, loc2] <- pairwise_adonis[[comp]]$R2[1]
  r2_matrix[loc2, loc1] <- pairwise_adonis[[comp]]$R2[1]  # Sim
}
diag(p_matrix) <- 1  # p-value = 1 
diag(r2_matrix) <- 0  # R² = 0 

# abbrevs in matrix
abbrev_locations <- location_abbrevs[locations]
names(abbrev_locations) <- NULL
dimnames(p_matrix) <- list(abbrev_locations, abbrev_locations)
dimnames(r2_matrix) <- list(abbrev_locations, abbrev_locations)

# Convert to long format
p_melt <- melt(p_matrix, na.rm = TRUE)
colnames(p_melt) <- c("Location1", "Location2", "p_value")
r2_melt <- melt(r2_matrix, na.rm = TRUE)
colnames(r2_melt) <- c("Location1", "Location2", "R2")
       
```

### Heatmap of adonis posthoc

```{r}
# Heatmap of p-values
heatmap_plot_p <- ggplot(p_melt, aes(x = Location1, y = Location2, fill = p_value)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient2(low = "white", mid = "white", high = "gray40",
                       midpoint = 0.025, limits = c(0, 1),
                       name = "p-value") +
  geom_text(aes(label = sprintf("%.3f", p_value)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "Pairwise PERMANOVA Bray Curtis", y = "Mangrove system") +
  theme(axis.text.x = element_blank(),  # Ocult axis x
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12),
        plot.title = element_text(size = 12))

# Heatmap de R²
heatmap_plot_r2 <- ggplot(r2_melt, aes(x = Location1, y = Location2, fill = R2)) +
  geom_tile(color = "gray40") +
  scale_fill_gradient(low = "white", high = "gray40",
                      name = "R²") +
  geom_text(aes(label = sprintf("%.3f", R2)), color = "black", size = 3) +
  theme_minimal() +
  labs(title = NULL,
       x = "Location", y = "Location") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 10))

# join cowplot
combined_heatmap <- plot_grid(
  heatmap_plot_p, heatmap_plot_r2,
  nrow = 2, align = "v", axis = "lr")

# show
combined_heatmap
```

## 04. Beta disper

::: callout-note
To validate the dissimilarity of the data, and to determine whether the differences found are due to changes in composition or to variations in dispersion within groups, the homogeneity of the dispersions between Mangrove type and Mangrove system was calculated using betadisper. Subsequently, the significance of the differences in dispersion was calculated with permutest
:::

```{r}
# library(vegan)
# library(phyloseq)
# Bray distance matrix
dist_matrix <- phyloseq::distance(ps, method = "bray")

```

### By Mangrove system

```{r}
# betadisper
disp_location <- betadisper(dist_matrix, sample_data$Mangrove_system)
anova_beta_disp <- anova(disp_location)

# Show dispersion by group
disp_location$group.distances
```

```{r}
# boxplot betadisper
beta_data <- data.frame(
  Group = metadata$Mangrove_system,
  Distance_to_centroid = disp_location$distances)
beta_plot <- ggplot(beta_data, aes(x = Group, y = Distance_to_centroid, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = loc_colors) +
  labs(title = "Betadisper by Mangrove system") +
  theme_classic()

#show
beta_plot
```

```{r}
ggsave("Figures/interior_impaired/Betadisper_location.pdf", beta_plot, bg='transparent', width = 5.5, height = 4)
```

Permutest

```{r}
# Significance permutest
perm_location <- permutest(disp_location, permutations = 999)
print(perm_location)
```

### By Depth

```{r}
# betadisper
disp_depth <- betadisper(dist_matrix, sample_data$Depths)
anova_beta_disp_depth <- anova(disp_depth)

# Show dispersion by group
disp_depth$group.distances
```

```{r}
# boxplot betadisper
beta_data_depth <- data.frame(
  Group = metadata$Depths,
  Distance_to_centroid = disp_depth$distances
)
beta_plot_depth <- ggplot(beta_data_depth, aes(x = Group, y = Distance_to_centroid, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = depth_colors) +
  labs(title = "Betadisper by Depths") +
  theme_classic()

#show
beta_plot_depth
```

```{r}
ggsave("Figures/interior_impaired/Betadisper_depth.pdf", beta_plot_depth, bg='transparent', width = 5.5, height = 4)
```

```{r}
# Significance permutest
perm_depth <- permutest(disp_depth, permutations = 999)
print(perm_depth)
```

## 05. **Plot with PERMANOVA, pairwise PERMANOVA and Betadisper**

```{r}
beta_text_ms <- paste("Betadisper Mangrove system, p-value:", 
                   round(anova_beta_disp$`Pr(>F)`[1], 3))
beta_text_depth <- paste("Betadisper Depth, p-value:", 
                   round(anova_beta_disp_depth$`Pr(>F)`[1], 3))
```

```{r}
# Add texts to NMDS
nmds <- nmds_plot_bray +
  annotate(
    "text", x = Inf, y = Inf, 
    label = paste(perm_text, beta_text_ms, beta_text_depth, sep = "\n"),
    hjust = 1.1, vjust = 1.2, size = 3, color = "black")

nmds
```

```{r}
ggsave("Figures/interior_impaired/NMDS.pdf", nmds, bg='transparent', width = 6, height = 4.5)
```

## Save rds beta diversity plot

```{r}
saveRDS(nmds, "rds/interior_impaired/beta-diversity.rds")
```

## Supplementary

```{r}
library(dplyr)

#extract permanova
tableS8_permanova <- as.data.frame(adonis_result) %>%
  mutate(Factor = rownames(.)) %>%
  dplyr::select(Factor, Df, SumOfSqs, R2, F.Model = F, p_value = `Pr(>F)`) %>%
  filter(!Factor %in% c("Total", "Residual")) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

# Extract PERMDISP
p_dispersion <- anova(disp_location)$`Pr(>F)`[1]

# join
tableS8_final <- tableS8_permanova %>%
  mutate(Dispersion_p = ifelse(Factor == "Mangrove_system", p_dispersion, NA)) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

#show
tableS8_final
#save
write.csv(tableS8_final, "Tables/interior_impaired/Supplementary_Table_S8_Interior_Impaired_BetaDisper.csv", row.names = FALSE)
```
