---
title: "Data Prepare"
---

::: callout-tip
## How are inland mangroves similar to a degraded coastal mangrove?❓

To answer this question, only samples that met the selection criteria were selected for comparison with each other. These samples were as follows:

1.  They hailed from the same depth range.
2.  They hailed from the strip, in the case of coastal mangroves.
3.  They hailed from the flooding season, in the case of coastal mangroves.
:::

## Load libraries

```{r}
#| message: false
#| warning: false
library(ggplot2)
library(qiime2R)
library(phyloseq)
library(dplyr)
library(microbiome)
```

## 01. Select samples

```{r}
#Create phyloseq objects with qiime2 artifacts
ps <- qza_to_phyloseq(
  features = "data/cluster_table_filter_freq218_emcEstero.qza",
  taxonomy = "data/taxonomyEstero.qza",
  metadata = "data/metadata14.tsv")
```

```{r}
# Extract metadata from phyloseq object
metadata <- data.frame(phyloseq::sample_data(ps), check.names = FALSE)
```

```{r}
# Filter Study zone and flood samples
filtered_data <- metadata %>%
  filter(season %in% c("flood")) %>%
  filter(Study_zone %in% c("Estero_pargo", "Rio San Pedro", "Laguna Cacahuate")) %>%  
  filter((Study_zone == "Estero_pargo" & Ecological_type == "Impaired") |  
         Study_zone == "Rio San Pedro" |  Study_zone == "Laguna Cacahuate")
```

```{r}
#Combine filtered metadata with phyloseq object to create a filtered phyloseq object

ps2 <- prune_samples(rownames(filtered_data), ps)
```

## 02. Data preparation

### 02.1 check data

```{r}
#library(microbiome)
microbiome::summarize_phyloseq(ps2)
```

### 02.2 Rename mangrove system

```{r}
# Rename specific levels in the Mangrove system column
#library(dplyr)
metadata_fltr <- filtered_data %>%
  mutate(`Mangrove_system` = case_when(
    Study_zone == "Rio San Pedro" ~ "San Pedro River",
    Study_zone == "Laguna Cacahuate" ~ "Fossil Lagoon",
    Study_zone == "Estero_pargo" ~ "Impaired Términos Lagoon",
    TRUE ~ Study_zone))
```

### 02.3 Add features

```{r}
# Extract sample_data
sample_data_df <- as(sample_data(ps2), "data.frame")

# Join sample_data with metadata to add Mangrove system
sample_data_df <- sample_data_df %>%
  left_join(dplyr::select(metadata_fltr, SampleID, Mangrove_system), by = "SampleID")
```

#### 02.3.1 Mangrove type

```{r}
# Create Mangrove_type
sample_data_df <- sample_data_df %>%
  mutate(Mangrove_type = case_when(
    `Mangrove_system` %in% c("Fossil Lagoon", "San Pedro River") ~ "Interior",
    `Mangrove_system` %in% "Impaired Términos Lagoon" ~ "Coastal",
    TRUE ~ NA_character_ ))
```

#### 02.3.2 Layers

```{r}
# Create layers
sample_data_df <- sample_data_df %>%
  mutate(Depths = case_when(
    depth %in% c("0-15", "5") ~ "Surface",
    depth %in% c("16-30", "20") ~ "Mid",
    depth %in% c("31-45","50-70", "40") ~ "Deep",
    TRUE ~ NA_character_))
```

#### 02.3.3 Save in phyloseq object

```{r}
# add .fastq.gz to SampleID 
sample_data_df$SampleID <- gsub("_R1$", "_R1.fastq.gz", sample_data_df$SampleID)

# add rownames as SampleID
rownames(sample_data_df) <- sample_data_df$SampleID

# update sample_data in phyloseq object
sample_data(ps2) <- sample_data(sample_data_df)

# check
tail(sample_data(ps2))

```

### 02.4 Create color pallete

```{r}
#Mangrove system colors
loc_colors <- c("Fossil Lagoon"= "#A7fcc1",
                "San Pedro River" = "#26B170",
                "Impaired Términos Lagoon" = "gray40")

#Mangrove type colors
mt_colors <- c("Interior"= "#A7fcc1",
                "Coastal" = "#329D9C")

#Depth colors
depth_colors <- c("Surface" = "#f5e5c4",
                 "Mid" = "#baa179",
                 "Deep" = "#80673b")
```

### 02.5 Check samples

```{r}
#| message: false
#| warning: false
# Samples Plot
metadata <- as(sample_data(ps2), "data.frame")

samples_plot <- ggplot(metadata, aes(x = Depths, y = ..count..,
fill = Depths)) + geom_bar(position = "dodge") +
  facet_wrap(~ Mangrove_system, scales = "free_x") +
  labs(title = "Samples of the analysis", y = "Number of Samples",
       x = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0)) + 
  scale_fill_manual(values = depth_colors)

# show plot
print(samples_plot)
```

### 02.6 Save objects in RDS to next analysis

```{r}
saveRDS(ps2, "rds/interior_impaired/physeq.rds")
```

