---
title: "1) Community assembly processes (βNTI / RCbray)"
---

# Community assembly processes

Which ecological processes — selection, dispersal, or drift — govern the assembly of inland relict sediment communities?❓

We use the phylogenetic null-model framework of Stegen et al., implemented in `iCAMP::qpen`, to partition pairwise community turnover into deterministic (selection) and stochastic (dispersal, drift) processes. **No ASVs were removed** prior to the analysis. βNTI is restricted to the inland relict sites, where a single primer-consistent dataset and a robust phylogenetic tree make the analysis well supported.

## Load libraries

```{r}
#| label: libs
#| message: false
#| warning: false
library(tidyverse)
```

------------------------------------------------------------------------

## The analysis (βNTI / RCbray with iCAMP::qpen)

### Import data and prepare matrices

```{r}
#| eval: false
#| echo: true
library(iCAMP); library(phyloseq); library(qiime2R)
set.seed(123)
nperm  <- 999      
ncores <- 4        

# Import (same pipeline as the rest of the site)
ps <- qza_to_phyloseq(
  features = "data/ASV_table_filter_freq218_emc.qza",
  tree     = "data/rooted-tree-iqtree.qza",
  taxonomy = "data/taxonomy.qza",
  metadata = "data/metadata2.tsv"
)

# Rename systems as in 03.1.RelativeAbundances
metadata <- as(sample_data(ps), "data.frame") %>%
  mutate(Location = case_when(
    Location == "El Cacahuate"    ~ "Fossil Lagoon",
    Location == "La Piedad"       ~ "Reforma Waterfalls",
    Location == "Dique Miguelito" ~ "Miguelito Dike",
    TRUE ~ Location))
sample_data(ps) <- sample_data(metadata)

# Community matrix (samples in rows) and phylogenetic distance — NO ASV filtering
comm <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) comm <- t(comm)
tree <- phy_tree(ps)
comunes <- intersect(colnames(comm), tree$tip.label)
comm <- comm[, comunes, drop = FALSE]
tree <- ape::drop.tip(tree, setdiff(tree$tip.label, comunes))
pd <- cophenetic(tree)[colnames(comm), colnames(comm)]
```

### Run qpen per site and pooled

```{r}
#| eval: false
#| echo: true
meta_df <- data.frame(sample_data(ps))
interiores <- c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike")

run_qpen <- function(samples, etiqueta) {
  cm  <- comm[rownames(comm) %in% samples, , drop = FALSE]
  cm  <- cm[, colSums(cm) > 0, drop = FALSE]
  pdi <- pd[colnames(cm), colnames(cm)]
  res <- qpen(comm = cm, pd = pdi, rand = nperm,
              nworker = ncores, sig.bNTI = 2, sig.rc = 0.95)
  res$result$grupo <- etiqueta
  res$result
}

resultados <- lapply(interiores, function(s)
  run_qpen(rownames(meta_df)[meta_df$Location == s], s))
resultados[["pooled"]] <- run_qpen(
  rownames(meta_df)[meta_df$Location %in% interiores], "All interiors (pooled)")

todos <- bind_rows(resultados)
dir.create("06.null-models-results", showWarnings = FALSE)
write_csv(todos, "06.null-models-results/qpen_pairwise_all.csv")

resumen <- todos %>%
  count(grupo, process) %>%
  group_by(grupo) %>%
  mutate(porcentaje = round(100 * n / sum(n), 1)) %>% ungroup()
write_csv(resumen, "06.null-models-results/qpen_process_summary.csv")
```

## Results and figures

### Project colour palette

```{r}
#| label: palette
# Localities — consistent with the rest of the site (loc_colors)
loc_colors <- c("Fossil Lagoon"          = "#A7fcc1",
                "Reforma Waterfalls"     = "#44C17B",
                "Miguelito Dike"         = "#2E8B57",
                "All interiors (pooled)" = "#b2ffa1")

# Assembly processes — greens/teals coherent with the site
process_colors <- c("Heterogeneous.Selection" = "#EFC7E6FF",#ffeeff
                    "Homogeneous.Selection"   = "#A790DBFF",#dcd3ff
                    "Dispersal.Limitation"    = "#D8D643FF",#FFFFDD
                    "Homogenizing.Dispersal"  = "#1BB6AFFF",#c4faf8
                    "Undominated"             = "gray80") #gray70

process_order <- c("Heterogeneous.Selection","Homogeneous.Selection",
                   "Dispersal.Limitation","Homogenizing.Dispersal","Undominated")
group_order   <- c("Fossil Lagoon","Reforma Waterfalls",
                   "Miguelito Dike","All interiors (pooled)")
```

### Load saved results

```{r}
#| label: load-results
#| message: false
todos   <- read_csv("06.null-models-results/qpen_pairwise_all.csv",
                    show_col_types = FALSE)
resumen <- read_csv("06.null-models-results/qpen_process_summary.csv",
                    show_col_types = FALSE)

todos$grupo     <- factor(todos$grupo, levels = group_order)
resumen$grupo   <- factor(resumen$grupo, levels = group_order)
resumen$process <- factor(resumen$process, levels = process_order)
```

### Fig 6A — βNTI distribution by group

```{r}
#| label: fig-bnti
#| fig-width: 7.5
#| fig-height: 5
#| warning: false
bnti_plot <- ggplot(todos, aes(grupo, bNTI, fill = grupo)) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.6) +
  scale_fill_manual(values = loc_colors) +
  labs(x = NULL, y = expression(beta*"NTI")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))

bnti_plot
```

> \|βNTI\| \> 2 indicates selection; βNTI \< −2 homogeneous selection, βNTI \> 2 heterogeneous selection. The dashed lines mark the ±2 thresholds.

```{r}
saveRDS(bnti_plot, "rds/assembly_process/bntiplot.rds")
```

### Fig 6B — Assembly process composition

```{r}
#| label: fig-stacked
#| fig-width: 8
#| fig-height: 5
raupplot <- ggplot(resumen, aes(grupo, porcentaje, fill = process)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = process_colors, name = "Assembly process") +
  labs(x = NULL, y = "% of pairwise comparisons") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(nrow = 2)) 

raupplot
```

```{r}
saveRDS(raupplot, "rds/assembly_process/rcplot.rds")
```

### Summary tables

```{r}
#| label: tables
#| message: false

bnti_site <- read_csv("06.null-models-results/bNTI_summary_by_site.csv",
                      show_col_types = FALSE)

knitr::kable(bnti_site, caption = "Mean βNTI per inland site.")
```

> **Key result.** Across all inland samples, deterministic selection dominated assembly (60.4% of comparisons; heterogeneous selection 50.4%), reproducibly across the three sites. Fossil Lagoon, the most isolated site, showed the largest contribution of dispersal limitation (mean βNTI 0.85), whereas the river sites were selection-dominated (mean βNTI 3.12 and 3.49). Fossil Lagoon differed from both river sites (Fisher, BH-adjusted *p* = 0.006 and 0.009), while the two river sites did not differ (*p* = 0.16).
