---
title: "1) Community assembly processes (βNTI / RCbray)"
---

# Community assembly processes

Which ecological processes — selection, dispersal, or drift — govern the assembly of inland relict sediment communities?❓

We use the phylogenetic null-model framework of Stegen et al., implemented in `iCAMP::qpen`, to partition pairwise community turnover into deterministic (selection) and stochastic (dispersal, drift) processes. **No ASVs were removed** prior to the analysis. βNTI is restricted to the inland relict sites, where a single primer-consistent dataset and a robust phylogenetic tree make the analysis well supported.

## Load libraries

```{r}
#| label: libs
#| message: false
#| warning: false
library(tidyverse)
```

------------------------------------------------------------------------

## The analysis (βNTI / RCbray with iCAMP::qpen)

### Import data and prepare matrices

```{r}
#| eval: false
#| echo: true
library(iCAMP); library(phyloseq); library(qiime2R)
set.seed(123)
nperm  <- 999      # randomizations
ncores <- 4        # bajado por OOM (fork overcommit con muchos workers)

# Import (same pipeline as the rest of the site)
ps <- qza_to_phyloseq(
  features = "data/ASV_table_filter_freq218_emc.qza",
  tree     = "data/rooted-tree-iqtree.qza",
  taxonomy = "data/taxonomy.qza",
  metadata = "data/metadata2.tsv"
)

# Rename systems as in 03.1.RelativeAbundances
metadata <- as(sample_data(ps), "data.frame") %>%
  mutate(Location = case_when(
    Location == "El Cacahuate"    ~ "Fossil Lagoon",
    Location == "La Piedad"       ~ "Reforma Waterfalls",
    Location == "Dique Miguelito" ~ "Miguelito Dike",
    TRUE ~ Location))
sample_data(ps) <- sample_data(metadata)

# Community matrix (samples in rows) and phylogenetic distance — NO ASV filtering
comm <- as(otu_table(ps), "matrix")
if (taxa_are_rows(ps)) comm <- t(comm)
tree <- phy_tree(ps)
comunes <- intersect(colnames(comm), tree$tip.label)
comm <- comm[, comunes, drop = FALSE]
tree <- ape::drop.tip(tree, setdiff(tree$tip.label, comunes))
pd <- cophenetic(tree)[colnames(comm), colnames(comm)]
```

### Run qpen per site and pooled

```{r}
#| eval: false
#| echo: true
meta_df <- data.frame(sample_data(ps))
interiores <- c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike")

run_qpen <- function(samples, etiqueta) {
  cm  <- comm[rownames(comm) %in% samples, , drop = FALSE]
  cm  <- cm[, colSums(cm) > 0, drop = FALSE]
  pdi <- pd[colnames(cm), colnames(cm)]
  res <- qpen(comm = cm, pd = pdi, rand = nperm,
              nworker = ncores, sig.bNTI = 2, sig.rc = 0.95)
  res$result$grupo <- etiqueta
  res$result
}

resultados <- lapply(interiores, function(s)
  run_qpen(rownames(meta_df)[meta_df$Location == s], s))
resultados[["pooled"]] <- run_qpen(
  rownames(meta_df)[meta_df$Location %in% interiores], "All interiors (pooled)")

todos <- bind_rows(resultados)
dir.create("06.null-models-results", showWarnings = FALSE)
write_csv(todos, "06.null-models-results/qpen_pairwise_all.csv")

resumen <- todos %>%
  count(grupo, process) %>%
  group_by(grupo) %>%
  mutate(porcentaje = round(100 * n / sum(n), 1)) %>% ungroup()
write_csv(resumen, "06.null-models-results/qpen_process_summary.csv")
```

### Significance of process differences between sites (Fisher's exact test)

```{r}
#| eval: false
#| echo: true
# βNTI values are non-independent pairwise comparisons, so differences in
# process composition between sites are tested with Fisher's exact test
# (Benjamini-Hochberg correction), NOT linear mixed models.
sitios <- c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike")
dat <- todos %>% filter(grupo %in% sitios)

tabla <- dat %>%
  count(grupo, process) %>%
  pivot_wider(names_from = process, values_from = n, values_fill = 0) %>%
  column_to_rownames("grupo") %>% as.matrix()

pares <- combn(sitios, 2, simplify = FALSE)
posthoc <- purrr::map_dfr(pares, function(pr) {
  sub <- tabla[pr, ]; sub <- sub[, colSums(sub) > 0, drop = FALSE]
  p <- fisher.test(sub, simulate.p.value = TRUE, B = 1e5)$p.value
  tibble(sitio_1 = pr[1], sitio_2 = pr[2], p_raw = p)
})
posthoc$p_adj <- p.adjust(posthoc$p_raw, method = "BH")
write_csv(posthoc, "06.null-models-results/posthoc_process_differences.csv")

resumen_bnti <- dat %>%
  group_by(grupo) %>%
  summarise(n_pares = n(),
            bNTI_medio = round(mean(bNTI), 2),
            bNTI_sd    = round(sd(bNTI), 2),
            pct_seleccion = round(100*mean(abs(bNTI) > 2), 1), .groups = "drop")
write_csv(resumen_bnti, "06.null-models-results/bNTI_summary_by_site.csv")
```

------------------------------------------------------------------------

```{r}
###############################################################################
#  SIGNIFICANCIA DE DIFERENCIAS EN PROCESOS DE ENSAMBLAJE ENTRE SITIOS
#  Test de PERMUTACIÓN A NIVEL DE MUESTRA (no Fisher)
#
#  Por qué: los valores de βNTI son comparaciones PAREADAS no independientes
#  (cada muestra participa en muchos pares). Fisher/chi2/LMM inflan la
#  significancia por pseudoreplicación. La permutación a nivel de muestra
#  respeta que la unidad independiente es la MUESTRA, no el par.
#
#  Lógica (estilo Stegen): se permutan las etiquetas de sitio sobre las
#  muestras; en cada permutación se reconstruye qué pares quedan "dentro" de
#  cada sitio y se recalcula el estadístico. La estructura pareada del βNTI
#  NO se toca; solo se baraja la pertenencia de cada muestra a su sitio.
###############################################################################

library(tidyverse)
set.seed(123)
nperm <- 9999   # permutaciones (9999 da p-values estables)

## ---------------------------------------------------------------------------
## 1. CARGAR los resultados pareados de qpen
## ---------------------------------------------------------------------------
# Necesitamos el archivo con TODOS los pares y su clasificación de proceso,
# que incluya a qué muestra (sample1, sample2) corresponde cada par.
# qpen$result trae columnas sample1, sample2, bNTI, RCbray, process.
pares <- read_csv("06.null-models-results/qpen_pairwise_all.csv",
                  show_col_types = FALSE)

# Nos quedamos SOLO con pares interiores (los 3 sitios), no el pooled
# (el pooled es un superset, no entra en la comparación entre sitios).
# Si 'grupo' marca el subconjunto, filtramos los tres sitios:
sitios <- c("Fossil Lagoon", "Reforma Waterfalls", "Miguelito Dike")

## ---------------------------------------------------------------------------
## 2. MAPA muestra -> sitio
## ---------------------------------------------------------------------------
# Necesitamos saber a qué sitio pertenece cada muestra. Lo tomamos del metadata.
# (ajustá la ruta/objeto si hace falta)
# meta debe tener: SampleID, Location
meta <- read_tsv("data/metadata2.tsv", show_col_types = FALSE)
# renombrar como en el resto del proyecto
meta <- meta %>%
  mutate(Location = case_when(
    Location == "El Cacahuate"    ~ "Fossil Lagoon",
    Location == "La Piedad"       ~ "Reforma Waterfalls",
    Location == "Dique Miguelito" ~ "Miguelito Dike",
    TRUE ~ Location)) %>%
  filter(Location %in% sitios)

# vector nombrado: nombre = SampleID, valor = sitio
col_id <- colnames(meta)[1]   # ajustá si el ID no es la 1a columna
sample2site <- setNames(meta$Location, meta[[col_id]])

## ---------------------------------------------------------------------------
## 3. PARES INTRA-SITIO observados
## ---------------------------------------------------------------------------
# Para comparar la composición de procesos ENTRE sitios, usamos los pares
# en los que AMBAS muestras pertenecen al mismo sitio (within-site pairs).
# Detectar columnas de muestra en el data frame de pares:
# qpen suele nombrarlas "name1"/"name2" o "sample1"/"sample2".
nm <- names(pares)
c1 <- nm[which(nm %in% c("sample1","name1","samp1","s1"))[1]]
c2 <- nm[which(nm %in% c("sample2","name2","samp2","s2"))[1]]
stopifnot(!is.na(c1), !is.na(c2))   # si falla, revisá los nombres reales

pares <- pares %>%
  mutate(site1 = sample2site[.data[[c1]]],
         site2 = sample2site[.data[[c2]]]) %>%
  filter(!is.na(site1), !is.na(site2))

intra <- pares %>% filter(site1 == site2) %>%
  mutate(site = site1)

## ---------------------------------------------------------------------------
## 4. ESTADÍSTICO OBSERVADO
## ---------------------------------------------------------------------------
# Resumimos cada sitio por su % de proceso determinístico (selección) o,
# más simple y robusto: el % de selección heterogénea (el proceso dominante).
# Definimos un estadístico de diferencia entre dos sitios = diferencia
# absoluta en la proporción de "Heterogeneous.Selection".

prop_proceso <- function(df, proc = "Heterogeneous.Selection") {
  df %>% group_by(site) %>%
    summarise(p = mean(process == proc), .groups = "drop")
}

obs <- prop_proceso(intra)

stat_pair <- function(tab, a, b) {
  abs(tab$p[tab$site == a] - tab$p[tab$site == b])
}

pares_sitios <- combn(sitios, 2, simplify = FALSE)
obs_stats <- map_dbl(pares_sitios, ~ stat_pair(obs, .x[1], .x[2]))
names(obs_stats) <- map_chr(pares_sitios, ~ paste(.x, collapse = " vs "))

## ---------------------------------------------------------------------------
## 5. PERMUTACIÓN A NIVEL DE MUESTRA
## ---------------------------------------------------------------------------
# Barajamos la etiqueta de sitio sobre las MUESTRAS, reconstruimos los pares
# intra-sitio bajo esa etiqueta permutada, y recalculamos el estadístico.
muestras   <- names(sample2site)
sitios_obs <- unname(sample2site)

perm_stats <- matrix(NA, nrow = nperm, ncol = length(obs_stats),
                     dimnames = list(NULL, names(obs_stats)))

for (i in seq_len(nperm)) {
  # barajar etiquetas de sitio entre muestras
  perm_site <- setNames(sample(sitios_obs), muestras)
  # reasignar sitio a cada muestra del par
  s1 <- perm_site[ intra[[c1]] ]
  s2 <- perm_site[ intra[[c2]] ]
  keep <- s1 == s2                      # pares que quedan intra-sitio
  tmp <- tibble(site = s1[keep], process = intra$process[keep])
  tabp <- tmp %>% group_by(site) %>%
    summarise(p = mean(process == "Heterogeneous.Selection"), .groups="drop")
  for (k in seq_along(pares_sitios)) {
    a <- pares_sitios[[k]][1]; b <- pares_sitios[[k]][2]
    pa <- tabp$p[tabp$site == a]; pb <- tabp$p[tabp$site == b]
    perm_stats[i, k] <- if (length(pa) && length(pb)) abs(pa - pb) else NA
  }
}

## ---------------------------------------------------------------------------
## 6. P-VALORES (proporción de permutaciones >= observado) + BH
## ---------------------------------------------------------------------------
p_raw <- map_dbl(seq_along(obs_stats), function(k) {
  nulldist <- perm_stats[, k]; nulldist <- nulldist[!is.na(nulldist)]
  (sum(nulldist >= obs_stats[k]) + 1) / (length(nulldist) + 1)  # +1 anti-cero
})
p_adj <- p.adjust(p_raw, method = "BH")

resultado <- tibble(
  comparison = names(obs_stats),
  obs_diff_hetSel = round(obs_stats, 3),
  p_perm = round(p_raw, 4),
  p_adj  = round(p_adj, 4)
)

print(resultado)
write_csv(resultado, "06.null-models-results/permutation_site_differences.csv")

## También guardamos las proporciones por sitio (para reportar descriptivo)
proporciones <- intra %>%
  count(site, process) %>%
  group_by(site) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>% ungroup()
write_csv(proporciones, "06.null-models-results/process_proportions_within_site.csv")

message("LISTO. permutation_site_differences.csv tiene los p-values honestos.")
###############################################################################
```

## Results and figures

### Project colour palette

```{r}
#| label: palette
# Localities — consistent with the rest of the site (loc_colors)
loc_colors <- c("Fossil Lagoon"          = "#A7fcc1",
                "Reforma Waterfalls"     = "#44C17B",
                "Miguelito Dike"         = "#2E8B57",
                "All interiors (pooled)" = "#b2ffa1")

# Assembly processes — greens/teals coherent with the site
process_colors <- c("Heterogeneous.Selection" = "#EFC7E6FF",#ffeeff
                    "Homogeneous.Selection"   = "#A790DBFF",#dcd3ff
                    "Dispersal.Limitation"    = "#D8D643FF",#FFFFDD
                    "Homogenizing.Dispersal"  = "#1BB6AFFF",#c4faf8
                    "Undominated"             = "gray80") #gray70

process_order <- c("Heterogeneous.Selection","Homogeneous.Selection",
                   "Dispersal.Limitation","Homogenizing.Dispersal","Undominated")
group_order   <- c("Fossil Lagoon","Reforma Waterfalls",
                   "Miguelito Dike","All interiors (pooled)")
```

### Load saved results

```{r}
#| label: load-results
#| message: false
todos   <- read_csv("06.null-models-results/qpen_pairwise_all.csv",
                    show_col_types = FALSE)
resumen <- read_csv("06.null-models-results/qpen_process_summary.csv",
                    show_col_types = FALSE)

todos$grupo     <- factor(todos$grupo, levels = group_order)
resumen$grupo   <- factor(resumen$grupo, levels = group_order)
resumen$process <- factor(resumen$process, levels = process_order)
```

### Fig 6A — βNTI distribution by group

```{r}
#| label: fig-bnti
#| fig-width: 7.5
#| fig-height: 5
#| warning: false
bnti_plot <- ggplot(todos, aes(grupo, bNTI, fill = grupo)) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", colour = "grey40") +
  geom_boxplot(width = 0.6, outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.6) +
  scale_fill_manual(values = loc_colors) +
  labs(x = NULL, y = expression(beta*"NTI")) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))

bnti_plot
```

> \|βNTI\| \> 2 indicates selection; βNTI \< −2 homogeneous selection, βNTI \> 2 heterogeneous selection. The dashed lines mark the ±2 thresholds.

```{r}
saveRDS(bnti_plot, "rds/assembly_process/bntiplot.rds")
```

### Fig 6B — Assembly process composition

```{r}
#| label: fig-stacked
#| fig-width: 8
#| fig-height: 5
raupplot <- ggplot(resumen, aes(grupo, porcentaje, fill = process)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = process_colors, name = "Assembly process") +
  labs(x = NULL, y = "% of pairwise comparisons") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(nrow = 2)) 

raupplot
```

```{r}
saveRDS(raupplot, "rds/assembly_process/rcplot.rds")
```

### Summary tables

```{r}
#| label: tables
#| message: false
posthoc <- read_csv("06.null-models-results/posthoc_process_differences.csv",
                    show_col_types = FALSE)
bnti_site <- read_csv("06.null-models-results/bNTI_summary_by_site.csv",
                      show_col_types = FALSE)

knitr::kable(bnti_site, caption = "Mean βNTI per inland site.")
knitr::kable(posthoc,  caption = "Fisher's exact test between sites (BH-adjusted).")
```

> **Key result.** Across all inland samples, deterministic selection dominated assembly (60.4% of comparisons; heterogeneous selection 50.4%), reproducibly across the three sites. Fossil Lagoon, the most isolated site, showed the largest contribution of dispersal limitation (mean βNTI 0.85), whereas the river sites were selection-dominated (mean βNTI 3.12 and 3.49). Fossil Lagoon differed from both river sites (Fisher, BH-adjusted *p* = 0.006 and 0.009), while the two river sites did not differ (*p* = 0.16).
