---
title: "Microbial association networks"
---

# Microbial association networks

Is the deterministic, depth-structured organization of inland relict communities also reflected in the architecture of microbial co-occurrence?❓

> **Placeholder.** This section will document the co-occurrence network analysis (ridge-regularized partial correlations on CLR-transformed top-200 taxa, balanced bootstrap resampling, fixed edge density 0.02, metrics computed with igraph, Louvain modularity, and depth-trend Spearman correlations). Content pending from collaborator. Code and rationale to be added here.

## Load libraries

```{r}
#| eval: false
# library(SpiecEasi)   # or the inference package used by the collaborator
# library(igraph)
# library(tidyverse)
```

## 01. Network inference

```{r}
#| eval: false
# TODO: collaborator code — ridge partial correlations, density 0.02,
#       balanced bootstrap (500 reps), top-200 taxa, CLR transform.
```

## 02. Network metrics

```{r}
#| eval: false
# TODO: largest connected component, modularity (Louvain), transitivity,
#       natural connectivity, robustness (random & targeted node removal).
```

## 03. Depth trend within inland sites

```{r}
#| eval: false
# TODO: Spearman correlations of metrics vs ordinal depth + Kruskal-Wallis.
```

> **Key results (to confirm with final figures):** connectivity, modularity and robustness decline with depth (ρ = −0.36, −0.38, −0.31, −0.41; all P \< 0.001); network structure is reproducible across the three inland sites (all bootstrap P \> 0.2); inland networks show higher transitivity than natural coastal networks (Δ = 0.27, P = 0.024).

## 04. Interior vs coastal comparisons

```{r}
#| eval: false
# TODO: bootstrap contrasts interior vs coastal (natural and impaired).
#       Only transitivity significant; report others as non-significant.
```
